### 2019-04-28 report/overview.tex
Déjà 11 pages ISSOU

\chapter{Overview of Optimal Transport}

\myminitoc

Le transport optimal permet de définir une géométrie et une distance sur l'ensemble des mesures de probabilités sur un espace.

\sect{Introduction}

\subs{Problème de Monge}

\begin{center}
	\begin{tikzpicture}[>=stealth']
		\draw[fill=brown!20] (-1, -1.8) rectangle (9, 0);
		\draw[fill=red, domain=0:2, smooth, variable=\x, fill opacity=0.9]
			plot({1.5*\x}, {\x * (27.21 + \x * (-126.6 + \x * (229.9 + \x * (-200.7
				+ \x * (84.24 - 13.61 * \x)))))}) -- (3.05, 0);
		\node[red] at (-0.2, 0.7) {$\mu$};
		\draw[fill=blue, fill opacity=0.9]
			(5, 0) -- (5.1, -1.2) -- (5.5, -1.2) -- (5.6, -1) -- (5.9, -1) --
			(6, -1.2) -- (6.4, -1.2) -- (6.5, -1) -- (6.8, -1) -- (6.9, -1.2) --
			(7.3, -1.2) -- (7.4, -1) -- (7.7, -1) -- (7.8, 0);
		\node[blue] at (8, -1.2) {$\nu$};
		\draw[ultra thick, purple] (0.4, 0) -- (0.4, 1.68);
		\draw[ultra thick, purple] (6.1, 0) -- (6.1, -1.2);
		\node[red] (x) at (0.4, -0.15) {$\bullet$};
		\node[blue] (y) at (6.1, 0.15) {$\bullet$};
		\node[below] at (x) {$x$};
		\node[above] at (y) {$y = T(x)$};
		\coordinate (y) at (6, 0.1);
		\draw[->, ultra thick, greenTikz, bend right=30]
			(x) edge node[below, black] {${\color{greenTikz}c}(x, T(x))$} (y);
	\end{tikzpicture}
\end{center}

\paragraph{Monge}
On se place dans un espace de probabilité $\Omega$. Puis on se donne $\mu$ et $\nu$, deux mesures de probabilités sur $\mathcal{P}(\Omega)$. De plus on se donne une fonction de coût $c : \Omega \times \Omega \rightarrow \R$. Cette fonction peut être la distance euclidienne par exemple. Le problème de Monge est alors le suivant :
$$ \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
\vspace{-2mm}
$$ \text{s.t.} \quad \forall A \in \mathcal{P}(\Omega), \, \nu(A) = \mu \left( T^{-1}(A) \right) $$
La fonction $T$ est appelé un plan de transport. Elle indique où est déplacé chaque morceau de masse. La masse qui se trouve en $x$ est déplacé à la position $T(x)$. La fonction $c$ permet alors de quantifier le coût du déplacement d'une unité de masse. On cherche alors à minimiser la distance moyenne de déplacement d'une unité de masse en transport optimal.

\paragraph{Limitations}
Le problème de cette formulation est qu'elle n'est pas utilisable sur des mesures discrètes. En effet si $\mu$ possède un Dirac plus grand que tous les Diracs de $\nu$ alors la construction d'un plan de transport devient impossible.

\subs{Problème de Kantorovich}

\begin{center}
	\begin{tikzpicture}[>={latex}, scale=1.2]
		\draw[->, greenTikz, ultra thick] (5.20, 3.65) -- node[black, above] {60} (3.96, 3.47);
		\fill[red] (5.20, 3.65) circle (0.136) node[white] {\tiny 60};
		\fill[blue] (3.65, 3.42) circle (0.313) node[white] {120};
		\draw[->, greenTikz, ultra thick] (6.69, 0.87) -- node[black, above right] {90} (5.21, 1.91);
		\fill[red] (6.69, 0.87) circle (0.221) node[white] {90};
		\fill[blue] (0.33, 0.18) circle (0.221) node[white] {90};
		\draw[->, greenTikz, ultra thick] (3.73, 0.65) -- node[black, left] {60} (3.66, 3.11);
		\draw[->, greenTikz, ultra thick] (3.73, 0.65) -- node[black, above] {90} (0.55, 0.21);
		\fill[red] (3.73, 0.65) circle (0.409) node[white] {150};
		\fill[blue] (5.03, 2.04) circle (0.221) node[white] {90};

		\node[red, below] at (5.2, 3.55) {\textbf{1}};
		\node[red, right=6] at (6.69, 0.87) {\textbf{2}};
		\node[red, below right=9] at (3.73, 0.65) {\textbf{3}};
		\node[blue, left=8] at (3.65, 3.42) {\textbf{A}};
		\node[blue, left=6] at (0.33, 0.18) {\textbf{B}};
		\node[blue, above=6] at (5.03, 2.04) {\textbf{C}};
	\end{tikzpicture}
\end{center}

\paragraph{Changement de plan de transport}
L'idée de Kantorovich a été de changer la fonction $T : \Omega \rightarrow \Omega$ en une probabilité $P$ sur l'espace produit $\mathcal{P}(\Omega \times \Omega)$. Sur l'exemple ci-dessus, la formulation de Monge nous interdit de séparé la masse de taille 150 en deux. Avec la formulation de Kantorovich on peut faire cela et ici une masse de taille 90 va a un endroit et une masse de taille 60 à un autre endroit. Dans le cas discret $P$ peut être représenté par une matrice :

\begin{center}
	\begin{tikzpicture}
		\draw[step=1, thick] (0, 0) grid (3, 3);
		\node at (0.5, 0.5) {\Large $p_{{\color{red}3}{\color{blue}A}}$};
		\node at (1.5, 0.5) {\Large $p_{{\color{red}3}{\color{blue}B}}$};
		\node at (2.5, 0.5) {\Large $p_{{\color{red}3}{\color{blue}C}}$};
		\node at (0.5, 1.5) {\Large $p_{{\color{red}2}{\color{blue}A}}$};
		\node at (1.5, 1.5) {\Large $p_{{\color{red}2}{\color{blue}B}}$};
		\node at (2.5, 1.5) {\Large $p_{{\color{red}2}{\color{blue}C}}$};
		\node at (0.5, 2.5) {\Large $p_{{\color{red}1}{\color{blue}A}}$};
		\node at (1.5, 2.5) {\Large $p_{{\color{red}1}{\color{blue}B}}$};
		\node at (2.5, 2.5) {\Large $p_{{\color{red}1}{\color{blue}C}}$};

		\node at (1.5, 3.5) {\large Matrice de Transport};
		\node[blue] at (0.5, -0.5) {\Large 120};
		\node[blue] at (1.5, -0.5) {\Large 90};
		\node[blue] at (2.5, -0.5) {\Large 90};
		\node[red] at (-0.5, 0.5) {\Large 60};
		\node[red] at (-0.5, 1.5) {\Large 90};
		\node[red] at (-0.5, 2.5) {\Large 150};

		\node at (7.5, 3.5) {\large Matrice des Distances};
		\draw[step=1, thick] (6, 0) grid (9, 3);
		\node at (6.5, 0.5) {\Large $d_{{\color{red}3}{\color{blue}A}}$};
		\node at (7.5, 0.5) {\Large $d_{{\color{red}3}{\color{blue}B}}$};
		\node at (8.5, 0.5) {\Large $d_{{\color{red}3}{\color{blue}C}}$};
		\node at (6.5, 1.5) {\Large $d_{{\color{red}2}{\color{blue}A}}$};
		\node at (7.5, 1.5) {\Large $d_{{\color{red}2}{\color{blue}B}}$};
		\node at (8.5, 1.5) {\Large $d_{{\color{red}2}{\color{blue}C}}$};
		\node at (6.5, 2.5) {\Large $d_{{\color{red}1}{\color{blue}A}}$};
		\node at (7.5, 2.5) {\Large $d_{{\color{red}1}{\color{blue}B}}$};
		\node at (8.5, 2.5) {\Large $d_{{\color{red}1}{\color{blue}C}}$};
	\end{tikzpicture}
\end{center}

\paragraph{Cas discret}
En notant $D$ notre matrice des distances, on obtient la formulation suivante dans le cas discret :
$$ \min_P \langle P, D \rangle = \sum_i \sum_j p_{ij} d_{ij} $$
\vspace{-4mm}
$$ \text{s.t.} \quad \left\{ \begin{array}{ll}
\forall i & \sum_j p_{ij} = \mu_i \\
\forall j & \sum_i p_{ij} = \nu_j
\end{array} \right. $$

\paragraph{Cas continue}
Dans le cas continue on introduit l'ensemble des probabilités conjointes suivant :
$$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega)~|~\forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$
Cet ensemble traduit les conditions du cas discret dans le cas continue. On reprend notre fonction de coût $c$ définie dans le problème de Monge. La formulation de Kantorovich devient le problème de minimisation suivant :
$$ \inf_{P \in \Pi(\mu, \nu)} \E_P \left[ c(X, Y) \right] = \int \int c(x, y) P(dx, dy) $$

\subs{Distance de Wasserstein}

Ce problème d'optimisation permet de définir une distance entre les mesures de probabilités.

\DEF{
	Étant donnée un espace de probabilité $\Omega$, une fonction de coût $c : \Omega \times \Omega \rightarrow \R$ et deux mesures de probabilités $\mu$ et $\nu$ dans $\mathcal{P}(\Omega)$, la \textbf{p-distance de Wasserstein} entre $\mu$ et $\nu$ est définie par :
	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x, y)^p P(dx, dy) \right)^{1 / p} $$
	\vspace{-3mm}
}

Cette distance peut permettre de calculer des barycentres entres des distributions de probabilité beaucoup plus naturels que des barycentres obtenus avec la norme $l_2$. La figure suivante nous montre la différence entre une interpolation (barycentre de deux distributions) obtenue via la distance de Wasserstein et une interpolation linéaire.

\begin{center}
	\begin{tikzpicture}[scale=0.85]
		\node at (5.5, 4.5) {\large Interpolation linéaire};
		\node[red] at (0.5, 2.3) {$\mu$};
		\node[blue] at (9.5, 2.7) {$\nu$};
		\fill[red] (0.000, 0.000) -- (0.050, 0.403) -- (0.100, 0.752) -- (0.150, 1.051) -- (0.200, 1.302) -- (0.250, 1.509) -- (0.300, 1.676) -- (0.350, 1.806) -- (0.400, 1.902) -- (0.450, 1.967) -- (0.500, 2.004) -- (0.550, 2.016) -- (0.600, 2.006) -- (0.650, 1.976) -- (0.700, 1.929) -- (0.750, 1.868) -- (0.800, 1.794) -- (0.850, 1.710) -- (0.900, 1.618) -- (0.950, 1.521) -- (1.000, 1.419) -- (1.050, 1.315) -- (1.100, 1.211) -- (1.150, 1.108) -- (1.200, 1.008) -- (1.250, 0.912) -- (1.300, 0.822) -- (1.350, 0.738) -- (1.400, 0.662) -- (1.450, 0.595) -- (1.500, 0.537) -- (1.550, 0.490) -- (1.600, 0.455) -- (1.650, 0.431) -- (1.700, 0.419) -- (1.750, 0.420) -- (1.800, 0.435) -- (1.850, 0.462) -- (1.900, 0.503) -- (1.950, 0.556) -- (2.000, 0.623) -- (2.050, 0.703) -- (2.100, 0.796) -- (2.150, 0.900) -- (2.200, 1.017) -- (2.250, 1.144) -- (2.300, 1.281) -- (2.350, 1.428) -- (2.400, 1.584) -- (2.450, 1.746) -- (2.500, 1.915) -- (2.550, 2.089) -- (2.600, 2.267) -- (2.650, 2.447) -- (2.700, 2.627) -- (2.750, 2.807) -- (2.800, 2.984) -- (2.850, 3.156) -- (2.900, 3.321) -- (2.950, 3.478) -- (3.000, 3.624) -- (3.050, 3.757) -- (3.100, 3.874) -- (3.150, 3.973) -- (3.200, 4.052) -- (3.250, 4.107) -- (3.300, 4.136) -- (3.350, 4.137) -- (3.400, 4.105) -- (3.450, 4.038) -- (3.500, 3.933) -- (3.550, 3.787) -- (3.600, 3.595) -- (3.650, 3.355) -- (3.700, 3.063) -- (3.750, 2.715) -- (3.800, 2.307) -- (3.850, 1.836) -- (3.900, 1.297) -- (3.950, 0.686) -- (4.000, 0.000);
		\fill[blue] (7.000, 0.000) -- (7.050, 0.215) -- (7.100, 0.418) -- (7.150, 0.611) -- (7.200, 0.792) -- (7.250, 0.964) -- (7.300, 1.126) -- (7.350, 1.278) -- (7.400, 1.421) -- (7.450, 1.555) -- (7.500, 1.681) -- (7.550, 1.798) -- (7.600, 1.907) -- (7.650, 2.008) -- (7.700, 2.102) -- (7.750, 2.188) -- (7.800, 2.268) -- (7.850, 2.341) -- (7.900, 2.408) -- (7.950, 2.468) -- (8.000, 2.522) -- (8.050, 2.571) -- (8.100, 2.614) -- (8.150, 2.652) -- (8.200, 2.685) -- (8.250, 2.713) -- (8.300, 2.737) -- (8.350, 2.755) -- (8.400, 2.770) -- (8.450, 2.781) -- (8.500, 2.787) -- (8.550, 2.790) -- (8.600, 2.790) -- (8.650, 2.785) -- (8.700, 2.778) -- (8.750, 2.767) -- (8.800, 2.754) -- (8.850, 2.737) -- (8.900, 2.718) -- (8.950, 2.696) -- (9.000, 2.672) -- (9.050, 2.645) -- (9.100, 2.615) -- (9.150, 2.584) -- (9.200, 2.550) -- (9.250, 2.514) -- (9.300, 2.476) -- (9.350, 2.436) -- (9.400, 2.394) -- (9.450, 2.350) -- (9.500, 2.305) -- (9.550, 2.257) -- (9.600, 2.208) -- (9.650, 2.157) -- (9.700, 2.104) -- (9.750, 2.050) -- (9.800, 1.993) -- (9.850, 1.935) -- (9.900, 1.875) -- (9.950, 1.814) -- (10.000, 1.750) -- (10.050, 1.685) -- (10.100, 1.618) -- (10.150, 1.549) -- (10.200, 1.477) -- (10.250, 1.404) -- (10.300, 1.329) -- (10.350, 1.251) -- (10.400, 1.172) -- (10.450, 1.090) -- (10.500, 1.005) -- (10.550, 0.918) -- (10.600, 0.828) -- (10.650, 0.736) -- (10.700, 0.640) -- (10.750, 0.542) -- (10.800, 0.440) -- (10.850, 0.336) -- (10.900, 0.227) -- (10.950, 0.116) -- (11.000, 0.000);
		\fill[purple] (0.000, 0.000) -- (0.050, 0.202) -- (0.100, 0.376) -- (0.150, 0.525) -- (0.200, 0.651) -- (0.250, 0.754) -- (0.300, 0.838) -- (0.350, 0.903) -- (0.400, 0.951) -- (0.450, 0.983) -- (0.500, 1.002) -- (0.550, 1.008) -- (0.600, 1.003) -- (0.650, 0.988) -- (0.700, 0.965) -- (0.750, 0.934) -- (0.800, 0.897) -- (0.850, 0.855) -- (0.900, 0.809) -- (0.950, 0.760) -- (1.000, 0.710) -- (1.050, 0.658) -- (1.100, 0.606) -- (1.150, 0.554) -- (1.200, 0.504) -- (1.250, 0.456) -- (1.300, 0.411) -- (1.350, 0.369) -- (1.400, 0.331) -- (1.450, 0.297) -- (1.500, 0.269) -- (1.550, 0.245) -- (1.600, 0.227) -- (1.650, 0.215) -- (1.700, 0.210) -- (1.750, 0.210) -- (1.800, 0.217) -- (1.850, 0.231) -- (1.900, 0.251) -- (1.950, 0.278) -- (2.000, 0.312) -- (2.050, 0.352) -- (2.100, 0.398) -- (2.150, 0.450) -- (2.200, 0.508) -- (2.250, 0.572) -- (2.300, 0.641) -- (2.350, 0.714) -- (2.400, 0.792) -- (2.450, 0.873) -- (2.500, 0.958) -- (2.550, 1.045) -- (2.600, 1.133) -- (2.650, 1.223) -- (2.700, 1.314) -- (2.750, 1.403) -- (2.800, 1.492) -- (2.850, 1.578) -- (2.900, 1.661) -- (2.950, 1.739) -- (3.000, 1.812) -- (3.050, 1.878) -- (3.100, 1.937) -- (3.150, 1.987) -- (3.200, 2.026) -- (3.250, 2.054) -- (3.300, 2.068) -- (3.350, 2.068) -- (3.400, 2.052) -- (3.450, 2.019) -- (3.500, 1.967) -- (3.550, 1.893) -- (3.600, 1.797) -- (3.650, 1.677) -- (3.700, 1.531) -- (3.750, 1.357) -- (3.800, 1.154) -- (3.850, 0.918) -- (3.900, 0.648) -- (3.950, 0.343) -- (4.000, 0.000);
		\fill[purple] (7.000, 0.000) -- (7.050, 0.107) -- (7.100, 0.209) -- (7.150, 0.305) -- (7.200, 0.396) -- (7.250, 0.482) -- (7.300, 0.563) -- (7.350, 0.639) -- (7.400, 0.711) -- (7.450, 0.778) -- (7.500, 0.840) -- (7.550, 0.899) -- (7.600, 0.953) -- (7.650, 1.004) -- (7.700, 1.051) -- (7.750, 1.094) -- (7.800, 1.134) -- (7.850, 1.171) -- (7.900, 1.204) -- (7.950, 1.234) -- (8.000, 1.261) -- (8.050, 1.286) -- (8.100, 1.307) -- (8.150, 1.326) -- (8.200, 1.343) -- (8.250, 1.357) -- (8.300, 1.368) -- (8.350, 1.378) -- (8.400, 1.385) -- (8.450, 1.390) -- (8.500, 1.394) -- (8.550, 1.395) -- (8.600, 1.395) -- (8.650, 1.393) -- (8.700, 1.389) -- (8.750, 1.384) -- (8.800, 1.377) -- (8.850, 1.369) -- (8.900, 1.359) -- (8.950, 1.348) -- (9.000, 1.336) -- (9.050, 1.322) -- (9.100, 1.308) -- (9.150, 1.292) -- (9.200, 1.275) -- (9.250, 1.257) -- (9.300, 1.238) -- (9.350, 1.218) -- (9.400, 1.197) -- (9.450, 1.175) -- (9.500, 1.152) -- (9.550, 1.129) -- (9.600, 1.104) -- (9.650, 1.078) -- (9.700, 1.052) -- (9.750, 1.025) -- (9.800, 0.997) -- (9.850, 0.968) -- (9.900, 0.938) -- (9.950, 0.907) -- (10.000, 0.875) -- (10.050, 0.842) -- (10.100, 0.809) -- (10.150, 0.774) -- (10.200, 0.739) -- (10.250, 0.702) -- (10.300, 0.664) -- (10.350, 0.626) -- (10.400, 0.586) -- (10.450, 0.545) -- (10.500, 0.502) -- (10.550, 0.459) -- (10.600, 0.414) -- (10.650, 0.368) -- (10.700, 0.320) -- (10.750, 0.271) -- (10.800, 0.220) -- (10.850, 0.168) -- (10.900, 0.114) -- (10.950, 0.058) -- (11.000, 0.000);
	\end{tikzpicture}
	\begin{tikzpicture}[scale=0.85]
		\node at (5.5, 4.5) {\large Interpolation de Wasserstein};
		\node[red] at (0.5, 2.3) {$\mu$};
		\node[blue] at (9.5, 2.7) {$\nu$};
		\fill[red] (0.000, 0.000) -- (0.050, 0.403) -- (0.100, 0.752) -- (0.150, 1.051) -- (0.200, 1.302) -- (0.250, 1.509) -- (0.300, 1.676) -- (0.350, 1.806) -- (0.400, 1.902) -- (0.450, 1.967) -- (0.500, 2.004) -- (0.550, 2.016) -- (0.600, 2.006) -- (0.650, 1.976) -- (0.700, 1.929) -- (0.750, 1.868) -- (0.800, 1.794) -- (0.850, 1.710) -- (0.900, 1.618) -- (0.950, 1.521) -- (1.000, 1.419) -- (1.050, 1.315) -- (1.100, 1.211) -- (1.150, 1.108) -- (1.200, 1.008) -- (1.250, 0.912) -- (1.300, 0.822) -- (1.350, 0.738) -- (1.400, 0.662) -- (1.450, 0.595) -- (1.500, 0.537) -- (1.550, 0.490) -- (1.600, 0.455) -- (1.650, 0.431) -- (1.700, 0.419) -- (1.750, 0.420) -- (1.800, 0.435) -- (1.850, 0.462) -- (1.900, 0.503) -- (1.950, 0.556) -- (2.000, 0.623) -- (2.050, 0.703) -- (2.100, 0.796) -- (2.150, 0.900) -- (2.200, 1.017) -- (2.250, 1.144) -- (2.300, 1.281) -- (2.350, 1.428) -- (2.400, 1.584) -- (2.450, 1.746) -- (2.500, 1.915) -- (2.550, 2.089) -- (2.600, 2.267) -- (2.650, 2.447) -- (2.700, 2.627) -- (2.750, 2.807) -- (2.800, 2.984) -- (2.850, 3.156) -- (2.900, 3.321) -- (2.950, 3.478) -- (3.000, 3.624) -- (3.050, 3.757) -- (3.100, 3.874) -- (3.150, 3.973) -- (3.200, 4.052) -- (3.250, 4.107) -- (3.300, 4.136) -- (3.350, 4.137) -- (3.400, 4.105) -- (3.450, 4.038) -- (3.500, 3.933) -- (3.550, 3.787) -- (3.600, 3.595) -- (3.650, 3.355) -- (3.700, 3.063) -- (3.750, 2.715) -- (3.800, 2.307) -- (3.850, 1.836) -- (3.900, 1.297) -- (3.950, 0.686) -- (4.000, 0.000);
		\fill[blue] (7.000, 0.000) -- (7.050, 0.215) -- (7.100, 0.418) -- (7.150, 0.611) -- (7.200, 0.792) -- (7.250, 0.964) -- (7.300, 1.126) -- (7.350, 1.278) -- (7.400, 1.421) -- (7.450, 1.555) -- (7.500, 1.681) -- (7.550, 1.798) -- (7.600, 1.907) -- (7.650, 2.008) -- (7.700, 2.102) -- (7.750, 2.188) -- (7.800, 2.268) -- (7.850, 2.341) -- (7.900, 2.408) -- (7.950, 2.468) -- (8.000, 2.522) -- (8.050, 2.571) -- (8.100, 2.614) -- (8.150, 2.652) -- (8.200, 2.685) -- (8.250, 2.713) -- (8.300, 2.737) -- (8.350, 2.755) -- (8.400, 2.770) -- (8.450, 2.781) -- (8.500, 2.787) -- (8.550, 2.790) -- (8.600, 2.790) -- (8.650, 2.785) -- (8.700, 2.778) -- (8.750, 2.767) -- (8.800, 2.754) -- (8.850, 2.737) -- (8.900, 2.718) -- (8.950, 2.696) -- (9.000, 2.672) -- (9.050, 2.645) -- (9.100, 2.615) -- (9.150, 2.584) -- (9.200, 2.550) -- (9.250, 2.514) -- (9.300, 2.476) -- (9.350, 2.436) -- (9.400, 2.394) -- (9.450, 2.350) -- (9.500, 2.305) -- (9.550, 2.257) -- (9.600, 2.208) -- (9.650, 2.157) -- (9.700, 2.104) -- (9.750, 2.050) -- (9.800, 1.993) -- (9.850, 1.935) -- (9.900, 1.875) -- (9.950, 1.814) -- (10.000, 1.750) -- (10.050, 1.685) -- (10.100, 1.618) -- (10.150, 1.549) -- (10.200, 1.477) -- (10.250, 1.404) -- (10.300, 1.329) -- (10.350, 1.251) -- (10.400, 1.172) -- (10.450, 1.090) -- (10.500, 1.005) -- (10.550, 0.918) -- (10.600, 0.828) -- (10.650, 0.736) -- (10.700, 0.640) -- (10.750, 0.542) -- (10.800, 0.440) -- (10.850, 0.336) -- (10.900, 0.227) -- (10.950, 0.116) -- (11.000, 0.000);
		\fill[purple] (3.500, 0.000) -- (3.550, 0.279) -- (3.600, 0.526) -- (3.650, 0.758) -- (3.700, 0.977) -- (3.750, 1.171) -- (3.800, 1.343) -- (3.850, 1.500) -- (3.900, 1.643) -- (3.950, 1.768) -- (4.000, 1.874) -- (4.050, 1.965) -- (4.100, 2.041) -- (4.150, 2.103) -- (4.200, 2.151) -- (4.250, 2.184) -- (4.300, 2.202) -- (4.350, 2.206) -- (4.400, 2.196) -- (4.450, 2.172) -- (4.500, 2.133) -- (4.550, 2.078) -- (4.600, 2.009) -- (4.650, 1.923) -- (4.700, 1.819) -- (4.750, 1.696) -- (4.800, 1.582) -- (4.850, 1.413) -- (4.900, 1.279) -- (4.950, 1.102) -- (5.000, 1.030) -- (5.050, 0.804) -- (5.100, 0.828) -- (5.150, 0.838) -- (5.200, 0.760) -- (5.250, 0.925) -- (5.300, 1.088) -- (5.350, 1.157) -- (5.400, 1.422) -- (5.450, 1.538) -- (5.500, 1.746) -- (5.550, 1.912) -- (5.600, 2.060) -- (5.650, 2.209) -- (5.700, 2.342) -- (5.750, 2.459) -- (5.800, 2.564) -- (5.850, 2.656) -- (5.900, 2.736) -- (5.950, 2.805) -- (6.000, 2.862) -- (6.050, 2.908) -- (6.100, 2.944) -- (6.150, 2.969) -- (6.200, 2.985) -- (6.250, 2.989) -- (6.300, 2.984) -- (6.350, 2.970) -- (6.400, 2.948) -- (6.450, 2.919) -- (6.500, 2.874) -- (6.550, 2.821) -- (6.600, 2.761) -- (6.650, 2.700) -- (6.700, 2.620) -- (6.750, 2.525) -- (6.800, 2.430) -- (6.850, 2.336) -- (6.900, 2.211) -- (6.950, 2.080) -- (7.000, 1.955) -- (7.050, 1.814) -- (7.100, 1.644) -- (7.150, 1.485) -- (7.200, 1.317) -- (7.250, 1.113) -- (7.300, 0.921) -- (7.350, 0.715) -- (7.400, 0.477) -- (7.450, 0.255) -- (7.500, 0.000);
	\end{tikzpicture}
\end{center}

On rappelle que le calcul du barycentre $\mu$ des des $N$ distributions $\nu_i$ associé aux poids $\lambda_i$ revient au problème de minimisation suivant :
$$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu, \nu_i) $$
Et voici maintenant ce que donne des barycentres avec 4 distributions :
\begin{center}
	\includegraphics[scale=0.4]{wasserstein_bar.png}
\end{center}

\sect{Comment calculer le transport optimal ?}

\subs{Dual}

Pour les calculs on revient vers les espaces discrets. Nos mesures de probabilité sont des sommes de diracs :
$$ \mu = \sum_{i=1}^n a_i \delta_{x_i} \qquad \nu = \sum_{j =1}^m b_j \delta_{y_j} $$
Notre fonction de coût est une matrice $D \in \R_+^{n \times m}$ tel que $D_{ij} = d(x_i, y_j)^p$. Notre distribution conjointe $P$ est aussi une matrice de $\R_+^{n \times m}$ qui doit vérifier certaines contraintes pouvant se représenter de manière matricielle. L'ensemble dans lequel se trouve $P$ est le suivant :
$$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbbm{1}_m = a, \, P^\trans \mathbbm{1}_n = b \right\} $$

\DEF{
	La définition de la distance de Wasserstein dans un espace discret est la suivante :
	$$ W_p^p(\mu, \nu) = \min_{P \in U(a, b)} \langle P, D \rangle $$
	\vspace{-6mm}
}

\paragraph{Dual} On s'intéresse maintenant au problème dual. On introduit le lagrangien :
$$ L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
La fonction objective du dual est :
$$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle $$
Si la matrice $D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans$ possède un coefficient négatif alors en faisant tendre le même coefficient de la matrice $P$ vers $+\infty$ on obtient pour minimum $-\infty$. En revanche si tous les coefficients sont positifs alors $\langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle$ est positif et il suffit de prendre $P=0$ pour se retrouver avec une valeur nulle. On obtient alors :
$$ g(\alpha, \beta) = \left\{ \begin{array}{ll}
\alpha^\trans a + \beta^\trans b & \text{si } D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \geqslant 0 \\
-\infty & \text{sinon}
\end{array}\right. $$
Finalement le dual est la maximisation de cette fonction $g$.

\DEF{
	Le \textbf{dual} du transport optimal a la formulation suivante :
	$$ W_p^p(\mu, \nu) = \max_{\alpha \in \R^n, \, \beta \in \R^m \atop \alpha_i + \beta_j \leqslant D(x_i, y_j)^p} \alpha^\trans a + \beta^\trans b $$
	\vspace{-5mm}
}

Avec un solveur de flot de coût minimum on peut résoudre cette optimisation linéaire en temps $\mathcal{O}(n^3 \log (n))$. Mais la solution est instable. Comme illustré sur la figure ci-dessous. Si on change un peu notre distance $D$ ou si on modifie un peu nos mesures source et destination $\mu$ et $\nu$ la valeur de $P$ peut subir un changement brutal puisque $P$ est un sommet de notre simplex. Une solution consiste à faire une régularisation de notre problème pour se retrouver avec un ensemble $U_\alpha(a, b)$ strictement convexe. Cet ensemble est strictement inclus dans $U(a, b)$ et la solution optimale de ce nouveau problème ne sera plus optimal pour le problème initial mais sera tout de même correcte. Finalement la régularisation que l'on va voire permet aussi d'obtenir un algorithme de résolution itératif de moindre complexité.

\begin{center}
	\begin{tikzpicture}[scale=20, >={latex}]
		\clip (0.2, -0.01) rectangle (0.51, 0.2);
		\draw[blue, thick] (0.1420, 0.1407) -- (0.5970, -0.1407);
		\draw[black, ->, thick] (0.2038, 0.1216) -- (0.2507, 0.1975);
		\fill[brown!40] (0.4503, 0.1444) -- (0.2251, 0.1444) -- (0.3695, 0.0000) -- (0.4503, -0.0000);
		\fill[greenTikz!60] (0.3827, 0.0946) -- (0.3827, 0.1104) -- (0.3804, 0.1148) -- (0.3759, 0.1206) -- (0.3692, 0.1263) -- (0.3602, 0.1307) -- (0.3556, 0.1321) -- (0.3489, 0.1336) -- (0.3264, 0.1336) -- (0.3219, 0.1321) -- (0.3151, 0.1278) -- (0.3129, 0.1234) -- (0.3129, 0.1191) -- (0.3151, 0.1133) -- (0.3241, 0.1018) -- (0.3444, 0.0888) -- (0.3534, 0.0859) -- (0.3714, 0.0859) -- (0.3804, 0.0917);
		\node[blue] at (0.3444, 0.0888) {$\bullet$};
		\node[blue, below=2] at (0.3444, 0.0888) {$P_\alpha$};
		\node[greenTikz] at (0.3489, 0.1119) {$\bullet$};
		\node[red] at (0.3695, 0.0000) {$\bullet$};
		\node[red, above=2] at (0.3695, 0) {$P$};
		\node at (0.482, 0.08) {$U(a, b)$};
		\node at (0.413, 0.12) {$U_\alpha(a, b)$};
		\node at (0.213, 0.165) {$D$};
	\end{tikzpicture}
\end{center}

\subs{Régularisation}

La régularisation que l'on effectue consiste à maximiser l'entropie de probabilité conjointe $P$. On fait apparaître un nouveau paramètre $\gamma$ qui contrôle cette régularisation et on obtient la distance suivante :

\DEF{
	La \textbf{distance de Wasserstein régularisée} est définie de la manière suivante :
	$$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$
	\vspace{-5mm}
}

\paragraph{Remarque}
On rappelle que l'on peut considérer deux variables aléatoires $X$ et $Y$ tel que $X \sim \mu$, $Y \sim \nu$ et $X, Y \sim P$. L'entropie de $P$ est maximale lorsque $X$ et $Y$ sont indépendants, c'est à dire lorsque $P = ab^\trans$ et l'entropie maximal est alors $H(\mu) + H(\nu)$. La régularisation donne donc une matrice avec des valeurs strictement positive plus réparties.

\begin{center}
	\includegraphics[scale=0.25]{reg_wass.png}
\end{center}

Par stricte convexité, il existe une unique matrice $P_\gamma$ qui minimise la distance :
$$ P_\gamma = \argmin_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$

\PROP{
	Il existe un unique couple de vecteurs $u$ et $v$ appartenant à $\R_+^n$ et $\R_+^m$ tel que :
	$$ P_\gamma = \text{diag}(u) K \text{diag}(v), \qquad K = e^{-D / \gamma} $$
	\vspace{-6mm}
}

\dem
On écrit le nouveau laplacien :
$$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
On calcule alors la dérivée partielle par rapport à $P_{ij}$ :
$$ \dfrac{\partial L}{\partial P_{ij}} = M_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$
Le dual est la maximisation du minimum du laplacien par rapport à $P$. Donc on doit avoir $P$ tel que la dérivée partielle est nulle :
$$ \dfrac{\partial L}{\partial P_{ij}} = 0 \Rightarrow P_{ij} = e^{\frac{\alpha_i}{\gamma}} e^{-\frac{D_{ij}}{\gamma}}  e^{\frac{\beta_j}{\gamma}} = u_i K_{ij} v_j $$
\findem

\paragraph{Sinkhorn}
On dispose alors d'une méthode itérative pour trouver cette matrice $P_\gamma$. On traduit la condition $P_\gamma \in U(a, b)$ :
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K \text{diag}(v) \mathbbm{1}_m & = & a \\
\text{diag}(v) K^\trans \text{diag}(u) \mathbbm{1}_n & = & b
\end{array} \right. $$
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
\text{diag}(u) K v & = & a \\
\text{diag}(v) K^\trans u & = & b
\end{array} \right. $$
On introduit alors $\odot$ le produit coefficient par coefficient :
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
u \odot K v & = & a \\
v \odot K^\trans u & = & b
\end{array} \right. $$
$$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrc}
u & = & a \, / \, K v \\
v & = & b \, / \, K^\trans u
\end{array} \right. $$
L'algorithme de Sinkhorn est alors le suivant :\\
\begin{algorithm}[H]
	\caption{Sinkhorn}
	\Repeat{convergence de $u, v$}{
		$u \gets a \, / \, K v$ \;
		$v \gets b \, / \, K^\trans u$ \;
	}
\end{algorithm}

\paragraph{Complexité}
Il a été prouvé que l'algorithme converge en temps linéaire. De plus une étape de calcul se fait en $\mathcal{O}(nm)$ mais les calculs peuvent être parallélisés. Il est aussi possible d'utiliser des convolutions sur une grille pour obtenir une complexité en $\mathcal{O}(n \log n)$.

\paragraph{Dual}
On peut aussi relever la formulation du dual avec régularisation :
$$ W_\gamma(\mu, \nu) = \max_{\alpha, \beta} \alpha^\trans a + \beta^\trans b - \gamma \left( e^{\alpha / \gamma} \right)^\trans K \left( e^{\beta / \gamma} \right) $$

\paragraph{Lien avec la divergence KL}
En posant le noyau :
$$ K_\gamma(i, j) = \exp \left( - \dfrac{d_{i, j}}{\gamma} \right) $$
On obtient une reformulation de la régularisation entropique avec la divergence KL :
$$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$

### 2019-04-28 report/papers.tex
Déjà 11 pages ISSOU

\chapter{Review of papers}

\myminitoc

\sect{Sinkhorn Distances: Lightspeed Computation of Optimal Transport}

\cite{Cut}

\sect{Wasserstein GAN}

\cite{Arj++}

\sect{Convolutional Wasserstein Distances: Efficient Optimal Transportation on Geometric Domains}

\cite{Goe++}

### 2019-04-28 report/overview.tex
Overview: outline for part III (applications of OT to ML problems...)

@@
 $$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$
+
+\sect{Optimal Transport and Machine Learning}
+
+\subs{Compute the optimal transport of what?}
+
+Since Sinkhorn can be parallelized, one can compute a large amount of Optimal
+Transport distances: for instance look into a database for similarity based
+retrieval, blablabla
+
+\paragraph{Color histograms}
+
+For image retrieval blablabla...
+
+\paragraph{Cloud of word}
+
+For similar documents retrieval blabla
+
+\subs{Wassersteinization}
+
+Introduce Wasserstein distance into optimizations problem.
+
+We show that we can computationaly differentiate, and therefore we can deal
+with it in a variety of optimization problem, including machine learning
+related opt. problems.
+
+Somehow revisiting machine learning problems.
+
+\paragraph{Averaging data}
+
+histograms, distribution, images, experiments (repetition of same expriment,
+MEG mesures, average over the experiments might be relevant)
+
+\paragraph{Aggregating distributions}
+
+From a very big dataset, split it in $J$ parts, learn the distribution (eg. w/
+Bayesaian Learning) from each of the part in parralel (on $J$ machines):
+the $J$ distributions should not be very different, but still the variance in
+the data will cause some variation: the idea is to use Wasserstein averaging
+to aggregate the distributions, and get a more meaningful one. Theoretic
+guaranties! :D
+
+Wasserstein posterior
+
+%\subs{Semi-suppervised learning setting}
+
+\paragraph{Semi-supervised learning setting}
+
+Wasserstein propagation: smooth a graph $(V,E)$ where an histogram is associated
+$\mu_v$ to each vertex $v$, with a susbset $S$ of the vertices that have a
+fixed histogram (known data).
+
+Solve the following optimization problem.
+
+$$
+    \underset{\substack{\mu_i \in \mathcal{P}\left(\Omega\right) \\ \text{for }i \in V\setminus S}}\min \sum_{(e_1, e_2) \in E} W_2^2\left(\mu_{e_1}, \mu_{e_2}\right)
+$$
+
+blabla
+
+\paragraph{omg dictionnary learning what is this????????? :oooooo}
+
+\subs{Wasserstein PCA, Geodesics and Inverse problem}
+
+Since the space of distributions with the OT geometry has negative curvature,
+we generalize the PCA (ie. projecting on one dimension spaces) into Generalized
+Principal Geodesics. We then want to explain stuff out of what we found;
+eg. affaire Bonneel vs nounours

### 2019-04-29 report/overview.tex
translation at the kilometer

@@
 
-Le transport optimal permet de définir une géométrie et une distance sur l'ensemble des mesures de probabilités sur un espace.
+Optimal transport aims to define a geometry and a distance on the set of
+mesures of probability on a space.
 
@@
 
-\subs{Problème de Monge}
+\subs{\textsc{Monge}'s Problem}
 
@@
 \paragraph{Monge}
-On se place dans un espace de probabilité $\Omega$.
-Puis on se donne $\mu$ et $\nu$, deux mesures de probabilités sur $\mathcal{P}(\Omega)$.
-De plus on se donne une fonction de coût $c : \Omega \times \Omega \rightarrow \R$.
-Cette fonction peut être la distance euclidienne par exemple.
-Le problème de Monge est alors le suivant :
+
+We work on a space of probability $\Omega$. We consider two mesures of
+probability $\mu$ and $\nu$ on $\mathcal{P}(\Omega)$. We also consider a cost function
+$c : \Omega \times \Omega \rightarrow \R$. This function can for instance be the Euclidean distance. Monge's problem is therefore the following
+
 $$ \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
@@
 $$ \text{s.t.} \quad \forall A \in \mathcal{P}(\Omega), \, \nu(A) = \mu \left( T^{-1}(A) \right) $$
-La fonction $T$ est appelé un plan de transport.
-Elle indique où est déplacé chaque morceau de masse.
-La masse qui se trouve en $x$ est déplacé à la position $T(x)$.
-La fonction $c$ permet alors de quantifier le coût du déplacement d'une unité de masse.
-On cherche alors à minimiser la distance moyenne de déplacement d'une unité de masse en transport optimal.
+
+The function $T$ is called a transport map. It carries the information
+regarding the displacement of each elementary unit of probability: the mass
+located in $x$ is moved toward $T(x)$. The function $c$ thus allows to
+quantify the cost of the displacement of one elementary unit of mass. We
+want to minimize the average distance travelled by mass units during the
+optimal transport.
 
 \paragraph{Limitations}
-Le problème de cette formulation est qu'elle n'est pas utilisable sur des mesures discrètes.
-En effet si $\mu$ possède un Dirac plus grand que tous les Diracs de $\nu$ alors la construction d'un plan de transport devient impossible.
 
-\subs{Problème de Kantorovich}
+The issue with this formulation of an optimal transport problem is that it
+doesn't handle discrete probability mesures. Indeed, if $\mu$ has a Dirac
+impulse bigger than every Dirac impulse of $\nu$, then the transport map
+cannot be build.
+
+\subs{Kantorovich's problem}
 
@@
 \paragraph{Changement de plan de transport}
-L'idée de Kantorovich a été de changer la fonction $T : \Omega \rightarrow \Omega$ en une probabilité $P$ sur l'espace produit $\mathcal{P}(\Omega \times \Omega)$.
-Sur l'exemple ci-dessus, la formulation de Monge nous interdit de séparé la masse de taille 150 en deux.
-Avec la formulation de Kantorovich on peut faire cela et ici une masse de taille 90 va a un endroit et une masse de taille 60 à un autre endroit.
-Dans le cas discret $P$ peut être représenté par une matrice :
+\paragraph{Change of transport plan}
+The idea of Kantorovich is to change the function $T : \Omega \rightarrow
+\Omega$ into a probability $P$ on the product space $\mathcal{P}(\Omega \times
+\Omega)$. On the upper example, Monge's formulation forbids to separate the
+masse of size 150 into two separate masses. With Kantorovich's formulation,
+this operation is accessible: the mass of $150$ is splitted into a mass of
+$90$ and a mass of $60$, both sent to different places. In the discrete
+case, we can represent what happens with a matrix.
 
@@
 
-		\node at (1.5, 3.5) {\large Matrice de Transport};
+		\node at (1.5, 3.5) {\large Transport Map};
 		\node[blue] at (0.5, -0.5) {\Large 120};
@@
 
-		\node at (7.5, 3.5) {\large Matrice des Distances};
+		\node at (7.5, 3.5) {\large Distance matrix};
 		\draw[step=1, thick] (6, 0) grid (9, 3);
@@
 
-\paragraph{Cas discret}
-En notant $D$ notre matrice des distances, on obtient la formulation suivante dans le cas discret :
+\paragraph{Discrete Case}
+
+If we note $D$ our disance matrix, we obtain the following formulation in the discrete case:
+
 $$ \min_P \langle P, D \rangle = \sum_i \sum_j p_{ij} d_{ij} $$
@@
 
-\paragraph{Cas continue}
-Dans le cas continue on introduit l'ensemble des probabilités conjointes suivant :
+\paragraph{Continuous Case}
+
+In the continuous setting, we introduce the following set of joint probabilities:
+
 $$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega)~|~\forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$
-Cet ensemble traduit les conditions du cas discret dans le cas continue.
-On reprend notre fonction de coût $c$ définie dans le problème de Monge.
-La formulation de Kantorovich devient le problème de minimisation suivant :
+
+This set encodes the conditions of the discrete case but in the very
+continuous setting. We consider once again our cost function $c$ defined
+in the Monge's problem. Kantorovich's formulation now becomes the following
+optimization problem:
+
 $$ \inf_{P \in \Pi(\mu, \nu)} \E_P \left[ c(X, Y) \right] = \int \int c(x, y) P(dx, dy) $$
+\subs{Wasserstein distance}
 
-\subs{Distance de Wasserstein}
-
-Ce problème d'optimisation permet de définir une distance entre les mesures de probabilités.
+This optimization problem allows us to define a distance between probability mesures:
 
 \DEF{
-	Étant donnée un espace de probabilité $\Omega$, une fonction de coût $c : \Omega \times \Omega \rightarrow \R$ et deux mesures de probabilités $\mu$ et $\nu$ dans $\mathcal{P}(\Omega)$, la \textbf{p-distance de Wasserstein} entre $\mu$ et $\nu$ est définie par :
-	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x, y)^p P(dx, dy) \right)^{1 / p} $$
+    Given a probability space $\Omega$, a cost function $c : \Omega \times
+    \Omega \rightarrow \R$ and two probabilitiy mesures $\mu$ and $\nu$ in
+    $\mathcal{P}(\Omega)$, the \textbf{p-Wasserstein-distance} between $\mu$
+    and $\nu$ is defined by :
+	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x,
+	y)^p P(dx, dy) \right)^{1 / p} $$
 	\vspace{-3mm}
@@
 
-Cette distance peut permettre de calculer des barycentres entres des distributions de probabilité beaucoup plus naturels que des barycentres obtenus avec la norme $l_2$.
-La figure suivante nous montre la différence entre une interpolation (barycentre de deux distributions) obtenue via la distance de Wasserstein et une interpolation linéaire.
+This distance allows to compute barycenters between several probability
+distributions, and obtain results that are way more natural than barycenters
+obtained with other norms, like $l_2$ for instance. The following
+figure compares an interpolation (progressive weighted barycenter of two
+distributions) obtained via Wasserstein-distance and a linear interpolation.
 
@@
 
-On rappelle que le calcul du barycentre $\mu$ des des $N$ distributions $\nu_i$ associé aux poids $\lambda_i$ revient au problème de minimisation suivant :
+We recall that the computation of the barycenter $\mu$ of the $N$ distributions
+$\nu_i$ associated to the weights $\lambda_i$ boils down to the following
+minimization problem.
 $$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu, \nu_i) $$
-Et voici maintenant ce que donne des barycentres avec 4 distributions :
+Now observe what the barycenter looks like with $4$ distributions:
 \begin{center}
@@
 
-\sect{Comment calculer le transport optimal ?}
+\sect{How to compute Optimal Transport?}
 
-\subs{Dual}
+\subs{Duality}
 
-Pour les calculs on revient vers les espaces discrets. Nos mesures de probabilité sont des sommes de diracs :
+To do the computations, we embrace the setting of discrete spaces. Our
+probability mesures are sum of diracs impulses:
+
 $$ \mu = \sum_{i=1}^n a_i \delta_{x_i} \qquad \nu = \sum_{j =1}^m b_j \delta_{y_j} $$
-Notre fonction de coût est une matrice $D \in \R_+^{n \times m}$ tel que $D_{ij} = d(x_i, y_j)^p$.
-Notre distribution conjointe $P$ est aussi une matrice de $\R_+^{n \times m}$ qui doit vérifier certaines contraintes pouvant se représenter de manière matricielle.
-L'ensemble dans lequel se trouve $P$ est le suivant :
+Our cost function is a matrix $D \in \R_+^{n \times m}$ such that $D_{ij} = d(x_i, y_j)^p$.
+Our joint distribution $P$ is also a matrix of $\R_+^{n \times m}$ that has to satisfy several constraints that can be represented matricially.
+The set which $P$ belongs to is:
+
 $$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbbm{1}_m = a, \, P^\trans \mathbbm{1}_n = b \right\} $$
@@
 \DEF{
-	La définition de la distance de Wasserstein dans un espace discret est la suivante :
+    The Wasserstein distance in a discrete space is defined as:
 	$$ W_p^p(\mu, \nu) = \min_{P \in U(a, b)} \langle P, D \rangle $$
@@
 
-\paragraph{Dual} On s'intéresse maintenant au problème dual. On introduit le lagrangien :
+\paragraph{Dual} We now focus on the dual problem. For this we introduce the lagrangian.
+
 $$ L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
-La fonction objective du dual est :
+The objective function of the dual is:
 $$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle $$
-Si la matrice $D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans$ possède un coefficient négatif alors en faisant tendre le même coefficient de la matrice $P$ vers $+\infty$ on obtient pour minimum $-\infty$.
-En revanche si tous les coefficients sont positifs alors $\langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle$ est positif et il suffit de prendre $P=0$ pour se retrouver avec une valeur nulle.
-On obtient alors :
+If the matrix $D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans$ has a negative coefficient, then by having this coefficient grow toward $+\infty$ we obtain a minimum of $-\infty$.
+But if all the coefficients are non negative, then  $\langle P, D - \alpha
+\mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle$ is non negative and we just have to take $P=0$ to obtain a non zero value. We therefore obtain:
 $$ g(\alpha, \beta) = \left\{ \begin{array}{ll}
@@
 \end{array}\right. $$
-Finalement le dual est la maximisation de cette fonction $g$.
+
+Finally, the dual problem is the maximization of this function $g$.
 
 \DEF{
-	Le \textbf{dual} du transport optimal a la formulation suivante :
+	The \textbf{dual} of the optimal transport problem is:
 	$$ W_p^p(\mu, \nu) = \max_{\alpha \in \R^n, \, \beta \in \R^m \atop \alpha_i + \beta_j \leqslant D(x_i, y_j)^p} \alpha^\trans a + \beta^\trans b $$
@@
 
-Avec un solveur de flot de coût minimum on peut résoudre cette optimisation linéaire en temps $\mathcal{O}(n^3 \log (n))$.
-Mais la solution est instable.
-Comme illustré sur la figure ci-dessous.
-Si on change un peu notre distance $D$ ou si on modifie un peu nos mesures source et destination $\mu$ et $\nu$ la valeur de $P$ peut subir un changement brutal puisque $P$ est un sommet de notre simplex.
-Une solution consiste à faire une régularisation de notre problème pour se retrouver avec un ensemble $U_\alpha(a, b)$ strictement convexe.
-Cet ensemble est strictement inclus dans $U(a, b)$ et la solution optimale de ce nouveau problème ne sera plus optimal pour le problème initial mais sera tout de même correcte.
-Finalement la régularisation que l'on va voire permet aussi d'obtenir un algorithme de résolution itératif de moindre complexité.
-
+Using a minimum flow solver, we can solve this linear optimization problem in
+$\mathcal{O}(n^3 \log (n))$. But the solution is not stable, as illustrated
+on the following figure. If we slightly move our distance $D$ or if we change
+the source and destination mesures, the value of $P$ can drastically change,
+since $P$ is a vertex of our simplex. One solution is to regularize the problem to fall back on a set $U_\alpha(a, b)$
+strictly convex. This set is strictly contained in $U(a, b)$ and the optimal solution of this new problem won't be optimal for the initial problem, but will remain valid.
+Finally, the regularization that we will describe allows to obtain an
+interative algorithm of small complexity.
 \begin{center}
@@
 
-\subs{Régularisation}
+\subs{Regularization}
 
-La régularisation que l'on effectue consiste à maximiser l'entropie de probabilité conjointe $P$.
-On fait apparaître un nouveau paramètre $\gamma$ qui contrôle cette régularisation et on obtient la distance suivante :
+The regularization that we choose consist in maximizing the entropy of the
+joint-probability $P$. We exhibit a new parameter $\gamma$ that controls
+the regularization and we obtain the following distance:
 
 \DEF{
-	La \textbf{distance de Wasserstein régularisée} est définie de la manière suivante :
+	The \textbf{regularized Wasserstein distance} is defined as follows:
 	$$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$
@@
 
-\paragraph{Remarque}
-On rappelle que l'on peut considérer deux variables aléatoires $X$ et $Y$ tel que $X \sim \mu$, $Y \sim \nu$ et $X, Y \sim P$.
-L'entropie de $P$ est maximale lorsque $X$ et $Y$ sont indépendants, c'est à dire lorsque $P = ab^\trans$ et l'entropie maximal est alors $H(\mu) + H(\nu)$.
-La régularisation donne donc une matrice avec des valeurs strictement positive plus réparties.
+\paragraph{Remark}
+Recall that we can consider two random variables $X$ and $Y$ such that $X \sim \mu$, $Y \sim \nu$ and $X,Y \sim P$.
+The entropy of $P$ is maximum when $X$ and $Y$ are independant, \emph{ie.} when $P = ab^\trans$ and this maximum entropy is thus $H(\mu) + H(\nu)$.
+The regularization therefore gives a matrix whose positive values are more
+distributed.
 
@@
 
-Par stricte convexité, il existe une unique matrice $P_\gamma$ qui minimise la distance :
+By strict convexity, there exist a unique matrix $P_\gamma$ that minimizes the distance:
 $$ P_\gamma = \argmin_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$
@@
 \PROP{
-	Il existe un unique couple de vecteurs $u$ et $v$ appartenant à $\R_+^n$ et $\R_+^m$ tel que :
+    There exist a unique couple of vectors $u$ and $v$ belonging to
+    $\R_+^n$ and $\R_+^m$ such that :
 	$$ P_\gamma = \text{diag}(u) K \text{diag}(v), \qquad K = e^{-D / \gamma} $$
@@
 \dem
-On écrit le nouveau laplacien :
+
+We write the new laplacian:
+
 $$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
-On calcule alors la dérivée partielle par rapport à $P_{ij}$ :
+We compute the partial derivative with respect to $P_{ij}$:
 $$ \dfrac{\partial L}{\partial P_{ij}} = M_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$
-Le dual est la maximisation du minimum du laplacien par rapport à $P$. Donc on doit avoir $P$ tel que la dérivée partielle est nulle :
+The dual is the maximization of the minimum of the laplacian with respect
+to $P$. We must look for $P$ such that the partial derivative is equal to zero:
 $$ \dfrac{\partial L}{\partial P_{ij}} = 0 \Rightarrow P_{ij} = e^{\frac{\alpha_i}{\gamma}} e^{-\frac{D_{ij}}{\gamma}}  e^{\frac{\beta_j}{\gamma}} = u_i K_{ij} v_j $$
@@
 \paragraph{Sinkhorn}
-On dispose alors d'une méthode itérative pour trouver cette matrice $P_\gamma$. On traduit la condition $P_\gamma \in U(a, b)$ :
+We have at our disposition an interative method to find this matrix $P_\gamma$. We use the condition $P_\gamma \in U(a,b)$:
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
@@
 \end{array} \right. $$
-On introduit alors $\odot$ le produit coefficient par coefficient :
+Then, we introduce $\odot$ the coefficient-wise product:
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
@@
 \end{array} \right. $$
-L'algorithme de Sinkhorn est alors le suivant :\\
+Sinkhorn's algorithm is thus the following: \\
 \begin{algorithm}[H]
 	\caption{Sinkhorn}
-	\Repeat{convergence de $u, v$}{
+	\Repeat{convergence of $u, v$}{
 		$u \gets a \, / \, K v$ \;
@@
 
-\paragraph{Complexité}
-Il a été prouvé que l'algorithme converge en temps linéaire.
-De plus une étape de calcul se fait en $\mathcal{O}(nm)$ mais les calculs peuvent être parallélisés.
-Il est aussi possible d'utiliser des convolutions sur une grille pour obtenir une complexité en $\mathcal{O}(n \log n)$.
+\paragraph{Complexity}
+It has been proved that the algorithm converges in linear time. Moreover,
+an iteration is done in $\mathcal{O}(mn)$ but the computation can be
+parallelized. It is also possible to use grid convolution to obtain a
+complexity of $\mathcal{O}(n \log n)$.
 
@@
 \paragraph{Lien avec la divergence KL}
-En posant le noyau :
+We write the kernel:
 $$ K_\gamma(i, j) = \exp \left( - \dfrac{d_{i, j}}{\gamma} \right) $$
-On obtient une reformulation de la régularisation entropique avec la divergence KL :
+We obtain a new formulation of the entropic regularization with the KL-divergence:
 $$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$

### 2019-04-29 report/overview.tex
Advance in overview

@@
 
+Now that we have convinced ourselves of the good mathematical foundations
+of Optimal Transport, and of the fact that we can actually compute it, we
+can wonder how we can put this tool to great use in the setting of machine
+learning.
+
 \subs{Compute the optimal transport of what?}
 
-Since Sinkhorn can be parallelized, one can compute a large amount of Optimal
-Transport distances: for instance look into a database for similarity based
-retrieval, blablabla
+In the previous section we have seen that we have an algorithm to compute
+optimal transport (Sinkhorn's algorithm), and that this algorithm can be
+parallelized. We can take advantage of this fact to actually compute optimal
+transport among huge databases. One of the application of this is similarity
+based retrieval.
 
@@
 
-For image retrieval blablabla...
+The color-histogram of an image is a tool that characterizes the color
+distribution of an image. Interestingly enough, it has been noticed that
+color-histogram are good tool to describe the similarity of two images. Since
+we now have at our disposal a tool to compute the distance between two
+distribution probability, we can apply it for histograms (that are really not
+that far from probability distribution, so much so that the algorithms and
+methods that we described earlier are relevant even without any fondamental
+change of setting). For a given image we can thus compute the distance between
+its histogram and the histogram of images comming from a bigger database:
+picking images that have the most similar histogram (in the sense of the
+Wasserstein distance) will therefore allow some form of similarity based
+image retrieval.
+
+\begin{center}
+	\includegraphics[scale=0.3]{image_histo.png}
+    \captionof{figure}{Two images and their respective color-histograms,
+    embedded in $\R^2$}
+\end{center}
 
@@
 
-For similar documents retrieval blabla
+We can apply the same technique as before, but instead of considering
+color-histograms, we can try to find some kind of distribution coming from
+other kind of documents, for example text documents. Some previous work,
+like the $\texttt{word2vec}$ framework allow for embedding of natural
+language words into an Euclidean space. A text therefore defines some sort
+of distribution of probability over this embedded space. As for images we
+can then compare this distribution to the distributions of other texts coming
+from a database (still in the sense of Wasserstein distance), and hopefully
+retrieve text dealing with similar topic, although there may be some
+significant differences in the actual vocabulary of those texts. We take
+advantage of the geometry of the distribution of probability in the embedded
+text, that tries to ensure that similar distributions are coming from texts
+dealing with similar topics.
 
@@
 
-Introduce Wasserstein distance into optimizations problem.
+We can go further than just using the Wasserstein to compute distance between
+distributions and use this as the main tool for the machine learning problems
+that we considered (we previously emphasized retrieval tasks) by introducing
+the Wasserstein distance on well studied machine learning settings and
+problems. We thus shed a new light on those problems, and the results are
+often fruitful. Some authors have coined the term $\textbf{Wassersteinization}$
+for this process.
 
-We show that we can computationaly differentiate, and therefore we can deal
-with it in a variety of optimization problem, including machine learning
-related opt. problems.
+\DEF{
+	$\textbf{Wassersteinization}$ is the process of introducing optimal
+	transport into an optimization or machine learning problem
+}
 
-Somehow revisiting machine learning problems.
+One way to approach this process is to introduce Wasserstein distance as a
+loss or fidelity term in optimization problems. This will motivate the interest
+toward the study of not only the very Wasserstein distance, but also of its
+derivative. More precisely we can show that we can computationaly evaluate
+the derivatives of the Wasserstein distance. Therefore we can deal with it a variety of optimization problems, including machine lerning related ones.
 
@@
 
-histograms, distribution, images, experiments (repetition of same expriment,
-MEG mesures, average over the experiments might be relevant)
+We can also see optimal transport as a normalization tool for data. One of
+the setting where this point of view as been proposed is the one of data
+averaging. We have already discussed earlier the benefits of Wasserstein's
+barycenters when compared to, say linear interpolations or $l_2$
+barycenters. One can see this put in good use in the case of data acquired
+through real world experiment. For instance, we can imagine a neurology
+setting, where the goal is to study the reaction of the brain of a patient
+to a certain precise stimulus. To do this we subjet the patient to the stimuli
+several time and we aim to submit the measures of the brain activity to
+a machine algorithm to study some of its features. But because of a whole
+lot of data acquisition hazards (noise, imprecisions in the measures, etc),
+the result will slighty differ from one experiment to another although it
+concerns the same stimulus. Therefore computing the average of the brain
+activity (that can easily seen to be analog to a probability distribution
+on the brain space), in the sense of Wasserstein barycenters can hopefully
+give access to more relevant data, without loosing the meaningful information.
+
+\begin{center}
+    \centering
+	\includegraphics[scale=0.3]{brain_map.png}
+    \captionof{figure}{On the left the average brain map among several
+    experiments, on the right the brain activity map on a single experiment.}
+\end{center}
+
+The same idea can be followed in other fields, like imagery for instance.
 
@@
 
-From a very big dataset, split it in $J$ parts, learn the distribution (eg. w/
-Bayesaian Learning) from each of the part in parralel (on $J$ machines):
-the $J$ distributions should not be very different, but still the variance in
-the data will cause some variation: the idea is to use Wasserstein averaging
-to aggregate the distributions, and get a more meaningful one. Theoretic
-guaranties! :D
-
-Wasserstein posterior
+Instead of processing the input of our machine learning algorithms, one can
+use this idea of aggregating several distribution on the output of machine
+learning algorithms. Imagine that we want to use Bayesian learning (\emph{ie.}
+learn a probability distribution, in a Bayesian framework), on a dataset so
+huge that it cannot even fit on a single machine, and would it fit that the
+computation time would be out of control. We can imagine to split the dataset
+into $J$ different parts and distribute it amongst $J$ machines running the
+same learning algorithms. The result of this process is a set of $J$
+distributions learned from each part of the dataset. Although they represent
+the same truth, because of the variation of the data those variation will
+be slightly different. Once again Wasserstein-averaging provides a useful
+tool to aggregate those distributions, into an hopefully meaningful one,
+closer the underlying real distribution. The good news is that this Wasserstein
+posterior aggregation method comes with theoretic statistical guaranties!
 

### 2019-04-29 report/papers.tex

@@
 
-\sect{Convolutional Wasserstein Distances: Efficient Optimal Transportation on Geometric Domains}
+\sect{Convolutional Wasserstein Distances: Efficient Optimal Transportation on Geometric Domains \cite{Goe++}}
 
-\cite{Goe++}
+Because a single step of Sinkhorn algorithm has a complexity in $\mathcal{O}(n^2)$, computing the Wassertein distance may take a while in many cases.
+That's why the researchers who write this paper thought about a more effective solution to compute a step.
+They found a way to use Gaussian convolution with a $\mathcal{O}(n \log n)$ complexity instead of a matrix-vector product.
+More generally they use a heat kernel.
+So instead of the convolution it can also be a sparse pre-factored linear system which can be computed in a complexity smaller than $\mathcal{O}(n^2)$.
+Furthermore by doing this, the convergence is still linear.
+Of course this can not be applied in all cases of optimal transport, but it can be used on very common geometric domains, like images or meshes.
+In addition of the computation of Wasserstein distance, they also propose a way to use convolution in the computation of Wasserstein barycentres and Wasserstein propagation.

### 2019-04-29 report/papers.tex
on continue la review

@@
 In addition of the computation of Wasserstein distance, they also propose a way to use convolution in the computation of Wasserstein barycentres and Wasserstein propagation.
+
+\paragraph{Using a heat kernel in the entropy-regularization}
+We recall what is the entropy-regularized Wasserstein distance between $\mu_0$ and $\mu_1$ on the domain $M$:
+$$ W_{2, \gamma}^2(\mu_0, \mu_1) = \inf_{\pi \in \Pi} \left[ \int \int_{M \times M} d(x, y)^2 \pi(x, y) dx dy \, - \, \gamma H(\pi) \right] = \gamma \left[ 1 + \min_{\pi \in \Pi} \text{KL}(\pi | \mathcal{K}_\gamma) \right]$$
+Where:
+$$ \Pi = \left\{ \pi \in \text{Prob}(M \times M) \, | \, \pi(\cdot, M) = \mu_0, \pi(M, \cdot) = \mu_1 \right\} $$
+$$ H(\pi) = - \int \int_{M \times M} \pi(x, y) \ln \pi(x, y) dx dy $$
+$$ \mathcal{K}_\gamma = e^{-d(x, y)^2 / \gamma} $$
+$$ \text{KL}(\pi | \mathcal{K}) = \int \int_{M \times M} \pi(x, y) \left[ \ln \dfrac{\pi(x, y)}{\mathcal{K}(x, y)} - 1 \right]\ln  dx dy $$
+This is a strictly convex problem thanks to the entropy.
+The idea of the paper is to use heat kernel because is some cases, the solution of the diffusion equation can be computed in a effective way.
+We denote by $\mathcal{H}_t(x, y)$ the diffusion between $x$ and $y$ after a time $t$ in the heat kernel.
+We have:
+$$ \mathcal{K}_\gamma \approx \mathcal{H}_{\gamma / 2}(x, y) $$
+We won't store $\mathcal{H}$ because it has a space complexity in $\mathcal{O}(n^2)$ and we want a lower complexity.
+But we generally know a way to apply $\mathcal{H}$ to a vector.
+In the case of images we apply Gaussian convolution with $\sigma^2 = \gamma$.
+\\
+In the case of triangle meshes we associate a weight to faces proportional to their area.
+We denote by $a$ the vector of these weights (In images of size $n \times m$ we set $a = 1 / (nm)$).
+Then we denote by $L$ the cotangent Laplacian and by $D_a$ the diagonal matrix with diagonal $a$.
+By discretizing the heat equation, we obtain:
+$$ w = \mathcal{H}_t(v) \Leftrightarrow \left( D_a + t L \right) w = v $$
+The linear system can be solved efficiently by pre-computing a sparse Cholesky factorization.
+
+\paragraph{Algorithm}
+We some computations we arrive at:
+
+\vspace{3mm}
+\begin{algorithm}[H]
+	\caption{\textsc{Convolutional-Sinkhorn}($\mu_0, \mu_1, H_t, a$)}
+	$v, w \gets 1$
+	\Repeat{convergence of $u, v$}{
+		$v \gets \mu_0 \oslash H_t(a \otimes w)$ \;
+		$v \gets \mu_1 \oslash H_t(a \otimes wv)$ \;
+	}
+	\Return $2t a^\trans \left[ \left( \mu_0 \otimes \ln v \right) + \left( \mu_1 \otimes \ln w \right) \right]$
+\end{algorithm}
+\vspace{3mm}
+
+Where $\oslash$ and $\otimes$ denote element-wise operations.
+As Sinkhorn algorithm we obtain a simple iterative algorithm where, this time, steps can be computed in $\mathcal{O}(n \log n)$ in many cases, where $n$ is the size of the domain $M$.
+
+\paragraph{Barycenter}
+An algorithm for computing barycenters is also provided.
+We won't enter into details but we give their algorithm for computing the barycenter of distributions $\mu_i$ associated with weights $\alpha_i$:
+
+\vspace{3mm}
+\begin{algorithm}[H]
+	\caption{\textsc{Convolutional-Barycenter}($\mu_0, \mu_1, H_t, a$)}
+	$v, w \gets 1$
+	\Repeat{convergence of $u, v$}{
+		$v \gets \mu_0 \oslash H_t(a \otimes w)$ \;
+		$v \gets \mu_1 \oslash H_t(a \otimes wv)$ \;
+	}
+	\Return $2t a^\trans \left[ \left( \mu_0 \otimes \ln v \right) + \left( \mu_1 \otimes \ln w \right) \right]$
+\end{algorithm}
+\vspace{3mm}

### 2019-04-29 report/papers.tex
Fin du premier article

@@
 	\caption{\textsc{Convolutional-Sinkhorn}($\mu_0, \mu_1, H_t, a$)}
-	$v, w \gets 1$
-	\Repeat{convergence of $u, v$}{
+	$v, w \gets 1$ \;
+	\Repeat{convergence of $v, w$}{
 		$v \gets \mu_0 \oslash H_t(a \otimes w)$ \;
-		$v \gets \mu_1 \oslash H_t(a \otimes wv)$ \;
+		$w \gets \mu_1 \oslash H_t(a \otimes v)$ \;
 	}
@@
 An algorithm for computing barycenters is also provided.
-We won't enter into details but we give their algorithm for computing the barycenter of distributions $\mu_i$ associated with weights $\alpha_i$:
+We won't enter into details but we give their algorithm for computing the barycenter of distributions $\mu_i$ associated with weights $\alpha_i$ for $1 \leqslant i \leqslant k$:
 
@@
 \begin{algorithm}[H]
-	\caption{\textsc{Convolutional-Barycenter}($\mu_0, \mu_1, H_t, a$)}
-	$v, w \gets 1$
-	\Repeat{convergence of $u, v$}{
-		$v \gets \mu_0 \oslash H_t(a \otimes w)$ \;
-		$v \gets \mu_1 \oslash H_t(a \otimes wv)$ \;
+	\caption{\textsc{Convolutional-Barycenter}($\{\mu_i\}, \{ \alpha_i \}, H_t, a$)}
+	$v_1, \dots, v_k \gets 1$ \;
+	$w_1, \dots, w_k \gets 1$ \;
+	\Repeat{convergence of $v_i, w_i$}{
+		$\mu \gets 1$ \;
+		\For{$i = 1, \dots, k$}{
+			$w_i \gets \mu_i \oslash H_t(a \otimes v_i)$ \;
+			$d_i \gets v_i \otimes H_t(a \otimes w_i)$ \;
+			$\mu \gets \mu \otimes d_i^{\alpha_i}$ \;
+		}
+		\vspace{3mm}
+		// Optional \;
+		$\mu \gets $ \textsc{Entropic-Sharpening}$\left( \mu, \max_i H(\mu_i) \right)$ \;
+		\vspace{3mm}
+		\For{$i = 1, \dots, k$}{
+			$v_i \gets v_i \otimes \mu \oslash d_i$ \;
+		}
 	}
@@
 \vspace{3mm}
+Where entropic sharpening allow us to make the entropy of the result $\mu$ smaller than the maximum entropy of all $\mu_i$.
+To do so if $H(\mu)$ is greater than $\max_i H(\mu_i)$, then we set $\mu$ to $\mu^\beta$ where $\beta$ is such that $H(\mu^\beta) = \max_i H(\mu_i)$.
+We can find $\beta$ using Newton's method.
+
+\paragraph{Applications}
+This has many applications in geometric domains. Applications that are given are:
+\begin{itemize}
+	\item Shape interpolation
+		\begin{center}
+		\includegraphics[scale=0.22]{shape_int.png}
+		\end{center}
+	\item BRDF design
+	\item Color histogram manipulation
+	\item Skeleton layout
+	\item Soft maps
+\end{itemize}
+Finally here are two images of barycenters they have computed.
+The first one is an image where we find barycenters of four images that are in the corners.
+The second is an interpolation between two cards.
+As we tried to reproduce it in the implementation part we put them here :
+
+\begin{center}
+	\includegraphics[scale=0.25]{bars_paper.png} \\ \vspace{3mm}
+	\includegraphics[scale=0.4]{cards_paper.png}
+\end{center}

### 2019-04-30 report/implementation.tex
début implémentation

\chapter{Implementation}

We decided to implement the last paper to compute Wasserstein distances on images in a efficient way. We then use convolutional Sinkhorn in SVM as Marco Cuturi used Sinkhorn in his paper to classify images of the MNIST digit database \cite{Cut}. The implementation of the convolutional Wassertein distance has been written in C++. We then write the Gramm matrix of a subset of the MNIST database in a file and use it in a Python script to use SVM from Sklearn library.

\paragraph{How to use our code}
OpenMP has been used to parallelize the code. If you don't have OpenMP already install, you can install it with \verb|sudo apt install libomp-dev|.
To compile the C++ code you just have to type \verb|make|. Then you need to obtain the train image datatset and the train label dataset of the MNIST database \url{http://yann.lecun.com/exdb/mnist/}. To do so you have to run the script \verb|get_mnist.sh| which will download and rename the files correctly. The executable \verb|main| created by \verb|make| command can be used in four ways.
\begin{itemize}
	\item  \verb|./main bar| to reproduce an image similar to the \figurename~\ref{barp}; An image containing different barycenters of four shapes.
	\item \verb|./main card| to reproduce images similar to the \figurename~\ref{cardp}; An interpolation of two images of cards.
	\item \verb|./main wass > kernel.txt| to compute a kernel with the convolutional Wasserstein distance of \verb|N_TRAIN| images of the MNIST dataset. We project \verb|N_TEST| other digit images of the MNIST dataset on the kernel space created. You can change the parameter \verb|N_TRAIN| and \verb|N_TEST| at the beginning of the file \verb|src/main.cpp|. The kernel will then be written if the file \verb|kernel.txt|.
	\item \verb|./main eucl > kernel.txt| to do the same as the previous command but with Euclidean distance instead of Wasserstein distance. 
\end{itemize}
Finally when a kernel file is computed you can test SVM on it thanks to the python script \verb|learning.py|. Because computing Wasserstein kernel may take a while we provide the Wasserstein kernel \verb|kernels/wass.txt| and the Euclidean kernel \verb|kernels/eucl.txt|. You can then run:
\begin{center}
	\verb=python3 learning.py < [kernels/wass.txt | kernels/eucl.txt]=
\end{center}
It will print the performances of these two kernels, show the projection of the data on the two principle components and finally show the kernel matrix.

\paragraph{Barycenters}
We first computed some barycenters to be sure that everything works. The code for computing Wasserstein distances and barycenters is in the file \verb|src/wasserstein.cpp|. The code that produce the barycenters that we will present is in the file \verb|examples.cpp|. The first example is the reproduction of \figurename~\ref{barp}. Here is what we obtained:

\begin{figure}[h]
	\centering
	\includegraphics[scale=0.5]{bars_own.png}
	\caption{Reproduction of \figurename~\ref{barp}}
\end{figure}

### 2019-04-30 learning.py

@@ -5,30 +5,38 @@ from sklearn.decomposition import PCA
 import numpy as np
 import pylab as pl
 
+# Load training set
 y_train = np.array(list(map(int, input().split())))
 n_train = len(y_train)
 D_train = []
 for _ in range(n_train):
 	D_train.append(list(map(float, input().split())))
 D_train = np.array(D_train)
+
+# Load test set
 y_test = np.array(list(map(int, input().split())))
 n_test = len(y_test)
 D_test = []
 for _ in range(n_test):
 	D_test.append(list(map(float, input().split())))
 D_test = np.array(D_test)
+
+# Substract the minimum distance
 mi = min(D_train.flatten())
 D_train = D_train - mi
 D_test = D_test - mi
 
+# Print the number of images of each classes
 ns = np.array([0]*10)
 for i in range(n_train):
 	ns[y_train[i]] += 1
 print(ns)
 
+# Gauss transformation
 def gauss(D, t):
 	return np.exp(- D**0.7 / t)
 
+# Use SVM with different parameters
 L = []
 for t in np.arange(3, 13, 0.5):
 	for C in np.arange(4.5, 11, 0.5):
@@ -36,14 +44,20 @@ for t in np.arange(3, 13, 0.5):
 		K_test = gauss(D_test, t)
 		svc = SVC(kernel='precomputed', C=C).fit(K_train, y_train)
 		L.append((t, C, svc.score(K_test, y_test), len(svc.support_)))
+
+# Print best results
 ms = max(a[2] for a in L)
 mt = 0
 for t, C, s, ls in L:
 	if s + 1.5 / n_test > ms:
-		print(round(t, 4), C, ls, round(s, 4))
+		print('t:', round(t, 4), 'C:', C, 'Number of support vectors:', ls, 'Result', str(round(100*s, 4)) + '%')
 	if s == ms:
 		mt = t
+print("===================================")
+print("Best Result:", str(round(100*ms, 4)) + '%')
+print("===================================")
 
+# Apply PCA on the kernel to visualize 
 su = np.array([[0.0]*n_train for _ in range(10)])
 K = gauss(D_train, mt)
 for i in range(n_train):
@@ -62,6 +76,8 @@ inds = np.array(inds).reshape(n_train, 1)
 K2 = K[inds, inds.T]
 pl.legend()
 pl.show()
+
+# Show kernel matrix
 pl.figure(figsize=(10, 10))
 pl.xlim(0, n_train)
 pl.ylim(n_train, 0)

### 2019-04-30 report/implementation.tex
fiigures
Début de la vitesse de la lumière

@@
 \end{figure}
+
+The result seems to be correct. We then try on larger images, the card images that we can find in the original paper presenting convolutional Wassertein \cite{Goe++}. Here is what we obtained:
+
+\begin{figure}[h]
+	\centering
+	\begin{tikzpicture}
+		\node[] at (0, 0) {\includegraphics[scale=0.3]{bar0.png}};
+		\node[] at (2.8, 0) {\includegraphics[scale=0.3]{bar3.png}};
+		\node[] at (5.6, 0) {\includegraphics[scale=0.3]{bar6.png}};
+		\node[] at (8.4, 0) {\includegraphics[scale=0.3]{bar8.png}};
+		\node[] at (11.2, 0) {\includegraphics[scale=0.3]{bar11.png}};
+		\node[] at (14, 0) {\includegraphics[scale=0.3]{bar14.png}};
+	\end{tikzpicture}
+	\caption{Reproduction of \figurename~\ref{cardp}}
+\end{figure}
+
+\paragraph{SVM usage}
+As proposed in the paper of Cuturi \cite{Cut}, we used the Wasserstein distance to compute the Gram matrix for 500 image and we used it as a kernel in a SVM.
+On a test set of 1500 images we obtained 92.4\% of good answers.
+Whereas with a kernel using Euclidean distance we obtain a good answer on only 88.4\%.
+To check these results you can run the python script \verb|learning.py| on the two kernels that are in the \verb|kernels| folder.

### 2019-04-30 report/overview.tex
Début de la vitesse de la lumière

@@
 
-\paragraph{Changement de plan de transport}
 \paragraph{Change of transport plan}
@@
 \begin{center}
-	\begin{tikzpicture}
+	\begin{tikzpicture}[scale=1.15]
 		\draw[step=1, thick] (0, 0) grid (3, 3);
@@
 	\begin{tikzpicture}[scale=0.85]
-		\node at (5.5, 4.5) {\large Interpolation linéaire};
+		\node at (5.5, 4.5) {\large Linear Interpolation};
 		\node[red] at (0.5, 2.3) {$\mu$};
@@
 	\begin{tikzpicture}[scale=0.85]
-		\node at (5.5, 4.5) {\large Interpolation de Wasserstein};
+		\node at (5.5, 4.5) {\large Wasserstein Interpolation};
 		\node[red] at (0.5, 2.3) {$\mu$};
@@
 $$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu, \nu_i) $$
-Now observe what the barycenter looks like with $4$ distributions:
-\begin{center}
+Now you can observe what the barycenter looks like with $4$ distributions in the \figurename~\ref{shp1}.
+\begin{figure}[h]
+	\centering
 	\includegraphics[scale=0.4]{wasserstein_bar.png}
-\end{center}
+	\captionsetup{justification=centering}
+	\caption{Each image contains several barycenters of the four shapes that are in the corners}
+	\label{shp1}
+\end{figure}
 
@@
 Sinkhorn's algorithm is thus the following: \\
-\begin{algorithm}[H]
+\begin{algorithm}
 	\caption{Sinkhorn}

### 2019-04-30 report/papers.tex
début implémentation
Début de la vitesse de la lumière

@@
 
-\sect{Sinkhorn Distances: Lightspeed Computation of Optimal Transport}
+\sect{Sinkhorn Distances: Lightspeed Computation of Optimal Transport \cite{Cut}}
 
-\cite{Cut}
+In this paper written in 2013, Cuturi proposed one of the most currently used algorithm to compute Wasserstein distances.
+He smoothed the classic optimal transport problem thanks to an entropy-regularization and it leads to a new distance which has a faster solver.
+He also showed that using regularization gives better results on MNIST classification problem than using the classic Wasserstein distance or obviously the Euclidean distance.
+This paper is part of the search for adequate and easily computable distances in machine learning.
+Wasserstein distance is better than Euclidean distance in many cases of machine learning but because the complexity to compute it is $\mathcal{O}(n^3 \log n)$ for $n$ the size of the data Wasserstein distance is rarely used.
+That's why techniques to compute the Wasserstein distance had to be develop.
+
+\paragraph{Idea}
+The main ideas of this paper are the use of an entropic-regularization to simplify computation when using Sinkhorn iterative algorithm and idea of computing the distances between one point and a family of other points in the same time, to use matrix-matrix multiplications instead of matrix-vector multiplications.
+Thus the algorithm can be implemented on GPU architectures.
+
+\paragraph{Regularization}
+We recall that the goal of the optimal transport problem in the Kantorovich formulation is to find a joint distribution $P \in \R^{d \times d}$ where the marginals equal the two distributions for which we want to compute the distance.
+That is to say $P \mathbbm{1}_n = a$ and $P^\trans \mathbbm{1}_n = b$ when we compute the distance between $a$ and $b$.
+We call $U(a, b)$ the set of all possible values for $P$.
+By denoting the entropy by $H$ we can show that:
+$$ H(P) \leqslant H(a) + H(b) $$
+And $H(P) = H(a) + H(b)$ when $P = a b^\trans$. We then restrict $P$ to the new set:
+$$ U_\alpha(a, b) = \{ P \in U(a, b) \, | \, \text{KL}(P || a b^\trans) \leqslant \alpha \} = \{ P \in U(a, b) \, | \, H(P) \geqslant H(a) + H(b) - \alpha \} $$
+Two reasons are given to this restriction.
+The first one is that it will be easier to compute the distance.
+The second is that without regularization $P$ has almost $2n-1$ non-zero coefficients.
+The transport is almost deterministic and it is not natural.
+That's why smoothing the transport plan with entropic regularization is presented as a good idea.
+In the paper, it is shown that this restriction induce a new distance:
+$$ d_{M, \alpha} = \min_{P \in U_\alpha(a, b)} \langle P, M \rangle $$
+Were $M$ is a distance matrix.
+
+\paragraph{Computation}
+By duality theory he obtained that to $\alpha$, corresponds a value $\lambda \geqslant 0$ such that the distance $d_{M, \alpha}$ is equal to the distance $d_M^\lambda$ defined by:
+$$ d_M^\lambda(a, b) = \langle P^\lambda, M \rangle, \quad \text{where } P^\lambda = \argmin_{P \in U(a, b)} \langle P, M \rangle - \dfrac{1}{\lambda} H(P) $$
+Then Cuturi show that this new distance can be computed with a much cheaper cost than the original Wassertein distance.
 
@@
 \begin{itemize}
-	\item Shape interpolation
-		\begin{center}
+	\item Shape interpolation (\figurename~\ref{shape})
+		\begin{figure}[h]
+			\centering
 			\includegraphics[scale=0.22]{shape_int.png}
-		\end{center}
+			\caption{Application to shape interpolation}
+			\label{shape}
+		\end{figure}
 	\item BRDF design
@@
 
-\begin{center}
-	\includegraphics[scale=0.25]{bars_paper.png} \\ \vspace{3mm}
-	\includegraphics[scale=0.4]{cards_paper.png}
-\end{center}
+\begin{figure}[h]
+	\centering
+	\includegraphics[scale=0.28]{bars_paper.png}
+	\caption{Barycenters of four images}
+	\label{barp}
+\end{figure}
+\begin{figure}[h]
+	\centering
+	\includegraphics[scale=0.45]{cards_paper.png}
+	\caption{Interpolation of two images of cards for times $t = 0, 0.25, 0.5, 0.75, 1$.}
+	\label{cardp}
+\end{figure}

### 2019-04-30 report/papers.tex
23 pages

@@
 $$ d_M^\lambda(a, b) = \langle P^\lambda, M \rangle, \quad \text{where } P^\lambda = \argmin_{P \in U(a, b)} \langle P, M \rangle - \dfrac{1}{\lambda} H(P) $$
-Then Cuturi show that this new distance can be computed with a much cheaper cost than the original Wassertein distance.
+Then Cuturi show that this new distance can be computed with a much cheaper cost than the original Wassertein distance $d_M$.
+In fact the iterative Sinkhorn algorithm presented below converge in few steps to the transport plan $P^\lambda$.
+The \figurename~\ref{recap1} summarize the relationships between all these distances.
 
-\sect{Wasserstein GAN}
+\vspace{3mm}
+\begin{algorithm}[H]
+	\caption{\textsc{Sinkhorn}$\left( a, \{ b_i \}_{i=1}^N, M, \lambda \right)$}
+	$B \gets \left[ b_1, \dots, b_N \right] \in \R^{n \times N}$ \;
+	$K \gets \exp \left( \lambda M \right)$ \;
+	$u = \left[ u_1, \dots, u_N \right] \gets \mathbbm{1}_{n \times N} \, / \, n$ \;
+	\Repeat{convergence of $u$}{
+		$u \gets a \, / \, \left( K \left( B \, / \, \left( K^\trans u \right) \right) \right)$ \;
+	}
+	$v = \left[ v_1, \dots, v_N \right] \gets B \, / \, \left( K^\trans u \right)$ \;
+	$P^\lambda_i \gets \text{diag}(u_i) K \text{diag}(v_i)$ \;
+	\Return $d = \left[ d_M^\lambda(a, b_1), \dots, d_M^\lambda(a, b_N) \right] = \left[ \left\langle M, P^\lambda_1 \right\rangle, \dots, \left\langle M, P^\lambda_N \right\rangle \right]$
+\end{algorithm}
+\vspace{3mm}
 
-\cite{Arj++}
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.5]{recap1.png}
+	\captionsetup{justification=centering}
+	\caption{Graphical representation of different distances and their relationships.}
+	\label{recap1}
+\end{figure}
+
+\paragraph{Results}
+The regularized Wasserstein distance is then used in a SVM to classifiy images of the MNIST database.
+The \figurename~\ref{comp} compare the error on a test set for different distances used in the SVM.
+The distance EMD is the classical optimal transport.
+As we can see the regularized version beat all distances.
+Furthermore he evaluated the execution time by stopping the iterations when $\| d_{i+1} / d_i - 1 \| < 10^{-4}$ where $d_i$ is the distance obtained at the end of the iteration $i$.
+He gave a comparison between EMD and regularized Wasserstein for different values of $\lambda$ in the \figurename~\ref{comp2}.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[scale=0.55]{comp.png}
+	\caption{Test error for different distances}
+	\label{comp}
+\end{figure}
+\begin{figure}[h]
+\centering
+\includegraphics[scale=0.5]{comp2.png}
+\caption{Execution time for different distances}
+\label{comp2}
+\end{figure}
+
+\newpage
+
+\sect{Wasserstein GAN \cite{Arj++}}
 
@@
 \end{itemize}
-Finally here are two images of barycenters they have computed.
+Finally \figurename~\ref{barp} and \figurename~\ref{cardp} are two images of barycenters they have computed.
 The first one is an image where we find barycenters of four images that are in the corners.
 The second is an interpolation between two cards.
-As we tried to reproduce it in the implementation part we put them here :
+As we tried to reproduce it in the implementation part we put them here:
 

### 2019-04-30 report/papers.tex
debut du gan

@@
 \sect{Wasserstein GAN \cite{Arj++}}
+
+This paper proposes a new GAN training algorithm based on Wasserstein distance.
+When we learn a probability we need to use a divergence between probability distributions.
+In a first time they show the advantages of the Wasserstein distance over all well-known other divergences on probabilities.
+Thanks to their WGAN (for Wasserstein GAN), they cure the problem of the need of maintaining a balance between training discriminator and training generator.
+They also observed that the mode dropping phenomenon is reduced.
+Furthermore theu GAN allow to plot the Wasserstein distance continuously which is very usefull for debugging and to have an estimation of the quality.
+
+\paragraph{Advantage of Wasserstein distance}
+They illustrate the interest of Wasserstein distance on the convergence of the probability $\Pp_\theta = g_\theta(Z) = (\theta, Z)$ toward $\Pp_0 = (0, Z)$ for $Z \sim U([0, 1])$ when $\theta$ tends to 0.
+We have:
+\begin{itemize}
+	\item $W(\Pp_0, \Pp_0\theta) = | \theta |$ where $W$ is the Wasserstein distance.
+	\item $JS(\Pp_0, \Pp_\theta) = \left\{ \begin{array}{ll}
+			\log 2 & \text{if } \theta \neq 0 \\
+			0 & \text{if } \theta = 0
+		\end{array} \right.$ where $JS$ is the Jensen-Shannon divergence.
+	\item $KL(\Pp_0, \Pp_\theta) = JS(\Pp_\theta, \Pp_0) = \left\{ \begin{array}{ll}
+			+ \infty & \text{if } \theta \neq 0 \\
+			0 & \text{if } \theta = 0
+		\end{array} \right.$ where $KL$ is the Kullback-Leibler divergence.
+	\item $\delta(\Pp_0, \Pp_\theta) = \left\{ \begin{array}{ll}
+			1 & \text{if } \theta \neq 0 \\
+			0 & \text{if } \theta = 0
+		\end{array} \right.$ where $\delta$ is the total variation distance defined as $\delta(\Pp_0, \Pp_1) = \sup_{A \in \Sigma} | \Pp_0(A) - \Pp_1(A) |$ with $\Sigma$ the set of all Borel subsets.
+\end{itemize}
+Thus the Wasserstein distance is the only one for which there is convergence.
+Then a gradient descent can be applied with the Wasserstein distance but it can't with other divergences.
+This a just an example but they also proved the following theorem which shows the continuity under some assumptions:
+
+\PROP{
+	tada
+}
 

### 2019-04-30 report/papers.tex
plus que les résultats WGAN

@@
 \begin{itemize}
-	\item $W(\Pp_0, \Pp_0\theta) = | \theta |$ where $W$ is the Wasserstein distance.
+	\item $W(\Pp_0, \Pp_\theta) = | \theta |$ where $W$ is the Wasserstein distance.
 	\item $JS(\Pp_0, \Pp_\theta) = \left\{ \begin{array}{ll}
@@
 \PROP{
-	tada
+	Let $\Pp_r$ be a distribution over $\mathcal{X}$ and let $Z$ be a distribution over another space $\mathcal{Z}$ (e.g Gaussian).
+Let $g : \R^d \times \mathcal{Z} \rightarrow \mathcal{X}$ be a function that will be denoted $g_\theta(z)$ for $g(\theta, z)$.
+We denote by $\Pp_\theta$ the distribution of $g_\theta(Z)$.
+Then:
+	\begin{itemize}
+		\item If $g$ is continuous in $\theta$, so is $W(\Pp_r, \Pp_\theta)$.
+		\item If $g$ is locally Lipschitz and satisfies the following regularity assumption: By denoting the local Lipschitz constant $L(\theta, z)$ we have $\E_{z \sim Z} \left[ L(\theta, z) \right] < + \infty$.
+Then $W(\Pp_r, \Pp_\theta)$ is continuous everywhere and almost differentiable everywhere.
+		\item The two previous statements are false for Jensen-Shannon and KL divergences.
+	\end{itemize}
 }
+
+They also give another theorem that well show that the Wasserstein distance is a very natural divergence for probability distributions:
+
+\PROP{
+	Consider a sequence of distributions $(\Pp_n)$ on a compact space $\mathcal{X}$.
+If $W(\Pp_n, \Pp) \rightarrow 0$ for a certain distribution $\Pp$, then $\Pp_n \xrightarrow{\mathcal{D}} \Pp$ where the limit is in the sense of the convergence in distribution.
+}
+
+\paragraph{WGAN}
+To construct their GAN they need to compute the gradient of $W(\Pp_r, \Pp_\theta)$ with respect to $\theta$. For that they consider the following dual formulation of the Wasserstein distance:
+$$ W(\Pp_r ,\Pp_\theta) = \max_{\| f \|_L \leqslant 1} \E_{x \sim \Pp_r} \left[ f(x) \right] - \E_{x \sim \Pp_\theta} \left[ f(x) \right] $$
+Where $\| f \|_L$ denote the Lipshitz coefficient of $f$. If $f$ is the solution to the maximization problem then the gradient is:
+$$ \nabla_\theta W(\Pp_r, \Pp_\theta) = - \E_{z \sim Z} \left[ \nabla_\theta f(g_\theta(z)) \right] $$
+To find $f$ they train a neural network parameterized with weights $w$ in a compact space $\mathcal{W}$ that lead to a function $f_w$.
+Then they can backprop $\theta$ using $\E_{z \sim Z} \left[ \nabla_\theta f_w (g_\theta(z)) \right]$.
+To have $w$ in a compact space, they use weight clamping of $w$.
+Their algorithm is in the \figurename~\ref{algo}.
+
+\begin{figure}[t]
+	\centering
+	\includegraphics[scale=0.6]{algorithm.png}
+	\caption{Algorithm of the WGAN}
+	\label{algo}
+\end{figure}
+
+\paragraph{results}
+First, the authors set up a small experiment to showcase the difference between GAN and WGAN (\figurename~\ref{compa}).
+There are two 1D Gaussian distributions, blue for real and green for fake.
+Train a GAN discriminator and WGAN critic to optimality, then plot their values over the space.
+The red curve is the GAN discriminator output, and the cyan curve is the WGAN critic output.
+
+\begin{figure}[h]
+	\centering
+	\includegraphics[scale=0.36]{wgan_res.png}
+	\caption{Comparison between a GAN discriminator and the WGAN Critic}
+	\label{compa}
+\end{figure}
+
+Both identify which distribution is real and which is fake, but the GAN discriminator does so in a way that makes gradients vanish over most of the space.
+In contrast, the weight clamping in WGAN gives a reasonably nice gradient over everything.
 

### 2019-04-30 report/overview.tex
final push

@@
     There exist a unique couple of vectors $u$ and $v$ belonging to
-    $\R_+^n$ and $\R_+^m$ such that :
+    $\R_+^n$ and $\R_+^m$ such that:
 	$$ P_\gamma = \text{diag}(u) K \text{diag}(v), \qquad K = e^{-D / \gamma} $$
@@
 
-blabla
-
-\paragraph{omg dictionnary learning what is this????????? :oooooo}
-
 \subs{Wasserstein PCA, Geodesics and Inverse problem}

### 2019-04-30 report/papers.tex
final push

@@
 This paper is part of the search for adequate and easily computable distances in machine learning.
-Wasserstein distance is better than Euclidean distance in many cases of machine learning but because the complexity to compute it is $\mathcal{O}(n^3 \log n)$ for $n$ the size of the data Wasserstein distance is rarely used.
+Wasserstein distance is better than Euclidean distance in many cases of machine learning but because the complexity to compute it is $\mathcal{O}(n^3 \log n)$ for $n$ the size of the data, Wasserstein distance is rarely used.
 That's why techniques to compute the Wasserstein distance had to be develop.
@@
 \paragraph{Idea}
-The main ideas of this paper are the use of an entropic-regularization to simplify computation when using Sinkhorn iterative algorithm and idea of computing the distances between one point and a family of other points in the same time, to use matrix-matrix multiplications instead of matrix-vector multiplications.
+The main ideas of this paper are the use of an entropic-regularization to simplify computation when using Sinkhorn iterative algorithm and the idea of computing the distances between one point and a family of other points in the same time, to use matrix-matrix multiplications instead of matrix-vector multiplications while computing distances between pair of points.
 Thus the algorithm can be implemented on GPU architectures.
@@
 \paragraph{Regularization}
-We recall that the goal of the optimal transport problem in the Kantorovich formulation is to find a joint distribution $P \in \R^{d \times d}$ where the marginals equal the two distributions for which we want to compute the distance.
+We recall that the goal of the optimal transport problem in the Kantorovich formulation is to find a joint distribution $P \in \R^{n \times n}$ where the marginals equal the two distributions for which we want to compute the distance.
 That is to say $P \mathbbm{1}_n = a$ and $P^\trans \mathbbm{1}_n = b$ when we compute the distance between $a$ and $b$.
@@
 The second is that without regularization $P$ has almost $2n-1$ non-zero coefficients.
-The transport is almost deterministic and it is not natural.
+Thus the transport is almost deterministic and it is not natural.
 That's why smoothing the transport plan with entropic regularization is presented as a good idea.
@@
 They also observed that the mode dropping phenomenon is reduced.
-Furthermore theu GAN allow to plot the Wasserstein distance continuously which is very usefull for debugging and to have an estimation of the quality.
+Furthermore their GAN allows to plot the Wasserstein distance continuously which is very usefull for debugging and to have an estimation of the quality.
 
@@
 			0 & \text{if } \theta = 0
-		\end{array} \right.$ where $\delta$ is the total variation distance defined as $\delta(\Pp_0, \Pp_1) = \sup_{A \in \Sigma} | \Pp_0(A) - \Pp_1(A) |$ with $\Sigma$ the set of all Borel subsets.
+		\end{array} \right.$ where $\delta$ is the total variation distance defined as \\ $\delta(\Pp_0, \Pp_1) = \sup_{A \in \Sigma} | \Pp_0(A) - \Pp_1(A) |$ with $\Sigma$ the set of all Borel subsets.
 \end{itemize}
 Thus the Wasserstein distance is the only one for which there is convergence.
-Then a gradient descent can be applied with the Wasserstein distance but it can't with other divergences.
-This a just an example but they also proved the following theorem which shows the continuity under some assumptions:
+Then a gradient descent can be applied with the Wasserstein distance but it can't be applied with other divergences.
+This is a just an example but they also proved the following theorem which shows the continuity under some assumptions:
 
@@
 Then they can backprop $\theta$ using $\E_{z \sim Z} \left[ \nabla_\theta f_w (g_\theta(z)) \right]$.
-To have $w$ in a compact space, they use weight clamping of $w$.
+To have $w$ in a compact space, they use weight clamping.
 Their algorithm is in the \figurename~\ref{algo}.
@@
 
+Another important point is that the quality of the generation is correlated to Wasserstein estimation as we can see in the \figurename~\ref{meaning}.
+We  can also see that the training of the critic function $f$ works well as the Wasserstein estimator strictly decrease.
+The gradient of the loss function is well computed.
+But this is not a method to compare the quality of different GANs as the estimator is defined with a constant scaling factor depending on the critic's architecture.
+They finally observed an improvement of the stability.
+
+\begin{figure}
+	\centering
+	\includegraphics[scale=0.45]{meaning.png}
+	\caption{Sample images at different iterations of the training}
+	\label{meaning}
+\end{figure}
+
 \sect{Convolutional Wasserstein Distances: Efficient Optimal Transportation on Geometric Domains \cite{Goe++}}
@@
 $$ \mathcal{K}_\gamma = e^{-d(x, y)^2 / \gamma} $$
-$$ \text{KL}(\pi | \mathcal{K}) = \int \int_{M \times M} \pi(x, y) \left[ \ln \dfrac{\pi(x, y)}{\mathcal{K}(x, y)} - 1 \right]\ln  dx dy $$
+$$ \text{KL}(\pi | \mathcal{K}) = \int \int_{M \times M} \pi(x, y) \left[ \ln \dfrac{\pi(x, y)}{\mathcal{K}(x, y)} - 1 \right] dx dy $$
 This is a strictly convex problem thanks to the entropy.
-The idea of the paper is to use heat kernel because is some cases, the solution of the diffusion equation can be computed in a effective way.
+The idea of the paper is to use heat kernel because in some cases, the solution of the diffusion equation can be computed in a effective way.
 We denote by $\mathcal{H}_t(x, y)$ the diffusion between $x$ and $y$ after a time $t$ in the heat kernel.
-We have:
+Just remember that the heat kernel is a solution of the heat equation:
+$$ \dfrac{\partial f}{\partial t} = \Delta f $$
+One can show the following approximation:
 $$ \mathcal{K}_\gamma \approx \mathcal{H}_{\gamma / 2}(x, y) $$
-We won't store $\mathcal{H}$ because it has a space complexity in $\mathcal{O}(n^2)$ and we want a lower complexity.
+We won't store $\mathcal{H}$ because it could have a space complexity in $\mathcal{O}(n^2)$ and we want a lower complexity.
 But we generally know a way to apply $\mathcal{H}$ to a vector.
@@
 \paragraph{Algorithm}
-We some computations we arrive at:
+With some computations we arrive at:
 

### 2019-10-27 report/implementation.tex
Rework some sentences and notations

@@
 
-We decided to implement the last paper to compute Wasserstein distances on images in a efficient way.
-We then use convolutional Sinkhorn in SVM as Marco Cuturi used Sinkhorn in his paper to classify images of the MNIST digit database \cite{Cut}.
-The implementation of the convolutional Wassertein distance has been written in C++.
-We then write the Gramm matrix of a subset of the MNIST database in a file and use it in a Python script to use SVM from Sklearn library.
+We decided to implement the last paper to compute Wasserstein distances
+on images in an efficient way. We then use convolutional Sinkhorn in an SVM
+classifier as Marco Cuturi used Sinkhorn in his paper to classify images of
+the MNIST digit database \cite{Cut}. The implementation of the convolutional
+Wassertein distance has been written in C++. We then write the Gramm matrix
+of a subset of the MNIST database in a file and use it in a Python script
+to use SVM from Sklearn library.
 
 \paragraph{How to use our code}
-OpenMP has been used to parallelize the code. If you don't have OpenMP already install, you can install it with \verb|sudo apt install libomp-dev|.
-To compile the C++ code you just have to type \verb|make|.
-Then you need to obtain the train image datatset and the train label dataset of the MNIST database \url{http://yann.lecun.com/exdb/mnist/}.
-To do so you have to run the script \verb|get_mnist.sh| which will download and rename the files correctly.
-The executable \verb|main| created by \verb|make| command can be used in four ways.
+OpenMP has been used to parallelize the code. If you don't have OpenMP already
+installed, you can install it with \verb|sudo apt install libomp-dev|.
+To compile the C++ code you just have to type \verb|make|. Then you need
+to obtain the train image datatset and the train label dataset of the MNIST
+database \url{http://yann.lecun.com/exdb/mnist/}. To do so you have to run
+the script \verb|get_mnist.sh| which will download and rename the files
+correctly. The executable \verb|main| created by \verb|make| command can be
+used in four ways.
+
 \begin{itemize}
-	\item  \verb|./main bar| to reproduce an image similar to the \figurename~\ref{barp}; An image containing different barycenters of four shapes.
-	\item \verb|./main card| to reproduce images similar to the \figurename~\ref{cardp}; An interpolation of two images of cards.
-	\item \verb|./main wass > kernel.txt| to compute a kernel with the convolutional Wasserstein distance of \verb|N_TRAIN| images of the MNIST dataset.
-We project \verb|N_TEST| other digit images of the MNIST dataset on the kernel space created.
-You can change the parameter \verb|N_TRAIN| and \verb|N_TEST| at the beginning of the file \verb|src/main.cpp|.
-The kernel will then be written if the file \verb|kernel.txt|.
-	\item \verb|./main eucl > kernel.txt| to do the same as the previous command but with Euclidean distance instead of Wasserstein distance. 
+	\item  \verb|./main bar| to reproduce an image similar to the
+	\figurename~\ref{barp}; An image containing different barycenters
+	of four shapes.
+
+	\item \verb|./main card| to reproduce images similar to the
+	\figurename~\ref{cardp}; An interpolation of two images of cards.
+
+	\item \verb|./main wass > kernel.txt| to compute a kernel with
+	the convolutional Wasserstein distance of \verb|N_TRAIN| images of
+	the MNIST dataset. We project \verb|N_TEST| other digit images of
+	the MNIST dataset on the kernel space created. You can change the
+	parameter \verb|N_TRAIN| and \verb|N_TEST| at the beginning of the
+	file \verb|src/main.cpp|. The kernel will then be written if the
+	file \verb|kernel.txt|.
+
+	\item \verb|./main eucl > kernel.txt| to do the same as the previous
+	command but with Euclidean distance instead of Wasserstein distance.
 \end{itemize}
-Finally when a kernel file is computed you can test SVM on it thanks to the python script \verb|learning.py|.
-Because computing Wasserstein kernel may take a while we provide the Wasserstein kernel \verb|kernels/wass.txt| and the Euclidean kernel \verb|kernels/eucl.txt|.
-You can then run:
+Finally when a kernel file is computed you can test SVM on it
+thanks to the python script \verb|learning.py|. Because computing
+Wasserstein kernel may take a while to compute we provide a pre-computed
+Wasserstein kernel \verb|kernels/wass.txt| as well as an Euclidean kernel
+\verb|kernels/eucl.txt|. You can then run:
 \begin{center}
@@
 \end{center}
-It will print the performances of these two kernels, show the projection of the data on the two principle components and finally show the kernel matrix.
+It will print the performances of these two kernels, show the projection of
+the data on the two principle components and finally show the kernel matrix.
 
 \paragraph{Barycenters}
-We first computed some barycenters to be sure that everything works.
-The code for computing Wasserstein distances and barycenters is in the file \verb|src/wasserstein.cpp|.
-The code that produce the barycenters that we will present is in the file \verb|examples.cpp|.
-The first example is the reproduction of \figurename~\ref{barp}.
-Here is what we obtained:
+We first computed some barycenters to be sure that everything works. The
+code for computing Wasserstein distances and barycenters is in the file
+\verb|src/wasserstein.cpp|. The code that produce the barycenters that we
+will present is in the file \verb|examples.cpp|. The first example is the
+reproduction of \figurename~\ref{barp}. Here is what we obtained:
 

### 2019-10-27 report/overview.tex
Rework some sentences and notations

@@
 
-Optimal transport aims to define a geometry and a distance on the set of
-mesures of probability on a space.
+Optimal Transport theory aims to define a geometry along with a notion of
+distance on the set of probability measures over a space.
 
@@
 \subs{\textsc{Monge}'s Problem}
+
+The history of optimal transport formulations can be traced back to Gaspard
+Monge's \emph{Mémoire sur la théorie des déblais et des remblais (1781)},
+studying how to move masses of sand or soil from one place to another. The
+main motivation of this \emph{Mémoire} being that we can assign a cost to
+the act of moving a massic particle from one place to another, and it is
+of particular interest to seek for the most optimal way of moving a global
+quantity of mass from one place to another.
 
@@
 
-\paragraph{Monge}
+\paragraph{Formalism of Monge's problem}
 
-We work on a space of probability $\Omega$. We consider two mesures of
-probability $\mu$ and $\nu$ on $\mathcal{P}(\Omega)$. We also consider a cost function
-$c : \Omega \times \Omega \rightarrow \R$. This function can for instance be the Euclidean distance. Monge's problem is therefore the following
+We work in a probability space $\Omega$. We consider two probability measures
+$\mu$ and $\nu$ on $\mathcal{P}(\Omega)$. We also consider a cost function
+$c : \Omega \times \Omega \rightarrow \R$. This function can for instance
+be the Euclidean distance. Monge's problem is therefore the following
 
-$$ \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
+$$ \min_{T: \Omega \to \Omega} \int_{\Omega} c \left( x, T(x) \right) \mu(x) $$
 \vspace{-2mm}
@@
 The function $T$ is called a transport map. It carries the information
-regarding the displacement of each elementary unit of probability: the mass
-located in $x$ is moved toward $T(x)$. The function $c$ thus allows to
-quantify the cost of the displacement of one elementary unit of mass. We
-want to minimize the average distance travelled by mass units during the
-optimal transport.
+regarding the displacement of each elementary unit of probability: the
+mass located in $x$ is moved toward $T(x)$. The function $c$ thus allows
+to quantify the cost of moving one elementary unit of mass. We want to
+minimize the average distance travelled by mass units during the optimal
+transport. The condition under which the optimization is performed ensures
+that all the mass of $\mu$ has been transported toward the distribution $\nu$.
 
@@
 
-The issue with this formulation of an optimal transport problem is that it
-doesn't handle discrete probability mesures. Indeed, if $\mu$ has a Dirac
-impulse bigger than every Dirac impulse of $\nu$, then the transport map
-cannot be build.
+The issue with such a formulation of an optimal transport problem is that
+it does not handle discrete probability measures. Indeed, if $\mu$ has a
+Dirac impulse bigger than every Dirac impulse of $\nu$, then the transport
+map cannot be build: we are only performing displacements of mass, and not
+actual repartition of mass.
 
 \subs{Kantorovich's problem}
+
+Kantorovich came up  with an alternative formulation for the optimal
+transport problem, aiming to handle discrete distributions.
 
@@
 \Omega$ into a probability $P$ on the product space $\mathcal{P}(\Omega \times
-\Omega)$. On the upper example, Monge's formulation forbids to separate the
+\Omega)$. On the above example, Monge's formulation forbids to split the
 masse of size 150 into two separate masses. With Kantorovich's formulation,
-this operation is accessible: the mass of $150$ is splitted into a mass of
-$90$ and a mass of $60$, both sent to different places. In the discrete
-case, we can represent what happens with a matrix.
+this operation becomes accessible: the mass of $150$ is splitted into a mass
+of $90$ and a mass of $60$, both sent to different places. In the discrete
+case, we can represent what happens through a matrix.
 
@@
 
-If we note $D$ our disance matrix, we obtain the following formulation in the discrete case:
+If we denote $D$ the distance matrix of the problem, we obtain the following
+formulation in the discrete case:
 
@@
 
+The two conditions encode the fact that we aim to move all the mass of $\mu$
+toward all the mass of $\nu$.
+
 \paragraph{Continuous Case}
 
-In the continuous setting, we introduce the following set of joint probabilities:
+In the continuous setting, we introduce the following set of joint
+probabilities, which encodes the analog of the two conditions in the discrete
+case, making sure that $\mu$ has been fully transported to $\nu$.
 
-$$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega)~|~\forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$
+$$ \Pi(\mu, \nu) = \left\{ P \in \mathcal{P}(\Omega \times \Omega) \mid \forall A, B \in \Omega, \, P(A \times \Omega) = \mu(A), \, P(\Omega \times B) = \nu(B) \right\} $$
 
-This set encodes the conditions of the discrete case but in the very
-continuous setting. We consider once again our cost function $c$ defined
-in the Monge's problem. Kantorovich's formulation now becomes the following
-optimization problem:
+We can rephrase the definition of $\Pi(\mu, \nu)$ as being the set of
+probability measures on $\Omega \times \Omega)$ with marginals $\mu$ and $\nu$.
+
+We consider once again a cost function $c$ defined as in the Monge's
+problem. Kantorovich's formulation now becomes the following optimization
+problem, where $X \sim \mu$ and $Y \sim \nu$:
 
 $$ \inf_{P \in \Pi(\mu, \nu)} \E_P \left[ c(X, Y) \right] = \int \int c(x, y) P(dx, dy) $$
+
+\paragraph{Existence of solutions} One can wonder whether solutions of
+Kantorovich's problem as formulated above exist. A rich litterature exists
+on this topic. For instance, in the case of measures on the $\mathbf{R}^n$
+space, with absolutely continuous measures $\mu$ and $\nu$ with respect to
+the Lebesgue measure, the existence (and uniqueness) of solutions will then
+depend on the properties of the cost function. If we consider the cost
+function $c_p(x,y) = | x - y |^p$, then we will have the following discussion:
+
+\begin{itemize}
+    \item $p > 1$: the strict convexity of $c_p$ ensures that there is a
+    unique solution to the Kantorovich problem
+    \item $p = 1$: here we can obtain the existence of a minimizer for the
+    Kantorovich, but we won't have the unicity of the solution
+    \item $p < 1$: in general, there will not exist solutions to the problem
+\end{itemize}
+
+More generally the existence of minimizers, as well as their uniqueness
+depends heavily on the structure of the space we consider, and the properties
+of both the measures and the cost function. A comprehensive study of those
+properties can be found in \cite{villani2003topics}.
+
 \subs{Wasserstein distance}
 
-This optimization problem allows us to define a distance between probability mesures:
+Having formulated this optimization problem allows us to define a distance
+between probability measures.
 
@@
     $\mathcal{P}(\Omega)$, the \textbf{p-Wasserstein-distance} between $\mu$
-    and $\nu$ is defined by :
+    and $\nu$ is defined as:
 	$$ W_p(\mu, \nu) = \left( \inf_{P \in \Pi(\mu, \nu)} \int \int c(x,
@@
 
-This distance allows to compute barycenters between several probability
-distributions, and obtain results that are way more natural than barycenters
-obtained with other norms, like $l_2$ for instance. The following
-figure compares an interpolation (progressive weighted barycenter of two
-distributions) obtained via Wasserstein-distance and a linear interpolation.
+This notion of distance allows to compute barycenters between several
+probability distributions, and obtain results that are way more natural
+than barycenters obtained with other norms, like the $l^2$ norm for
+instance. The following figure compares an interpolation (\emph{i.e.}
+a progressive weighted barycenter of two distributions) obtained under
+Wasserstein-distance and a linear interpolation (the two target distributions
+being the blue and the red one, and the purple shape being an intermediate
+interpolation).
 
@@
 
-We recall that the computation of the barycenter $\mu$ of the $N$ distributions
-$\nu_i$ associated to the weights $\lambda_i$ boils down to the following
-minimization problem.
-$$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu, \nu_i) $$
-Now you can observe what the barycenter looks like with $4$ distributions in the \figurename~\ref{shp1}.
+What emerges from those two interpolations is that the Wasserstein distance
+seems particularily well suited to manipulate distributions, as it appears
+to take into account the specificities (like modes, shape, etc) of both
+the source and the target distribution, and the resulting barycenter looks
+like a distribution combining those into one distribution, while the $l^2$
+barycenter does not merge those specificities into one distribution but
+rather renormalize the source and target distributions.
+
+We can extend this idea to the computation of the barycenter of an arbitrary
+number of distributions. Formally, the computing the barycenter $\mu$ of $N$
+distributions $\nu_i$ weighted by the $\lambda_i$ boils down to the following
+minimization problem:
+
+$$ \min_{\mu \in \mathcal{P}(\Omega)} \sum_{i=1}^N \lambda_i W_p^p(\mu,
+\nu_i) $$
+
+Now we can observe what the barycenter looks like with $4$ distributions
+(seen as images) on $\mathbf{R}^2$ in the \figurename~\ref{shp1}.
+
 \begin{figure}[h]
-	\centering
-	\includegraphics[scale=0.4]{wasserstein_bar.png}
-	\captionsetup{justification=centering}
-	\caption{Each image contains several barycenters of the four shapes that are in the corners}
+	\centering \includegraphics[scale=0.4]{wasserstein_bar.png}
+	\captionsetup{justification=centering} \caption{Each image contains
+	several barycenters of the four shapes that are in the corners}
 	\label{shp1}
@@
 
+Now that we have a convenient formulation of the optimal transport problem,
+we can study how to effectively compute minimizers of the optimal transport
+problem. For this we will introduce several new formulation, more adapted
+to effective solving. Sometimes, it is worth noting that we are only interested
+in computing the Wasserstein distance between two distributions, rather than
+the actual transport map.
+
 \subs{Duality}
 
-To do the computations, we embrace the setting of discrete spaces. Our
-probability mesures are sum of diracs impulses:
+To effectively solve the optimal transport problem, we embrace the setting
+of discrete spaces. In this setting, probability measures can be seen as a
+sum of diracs impulses.
 
 $$ \mu = \sum_{i=1}^n a_i \delta_{x_i} \qquad \nu = \sum_{j =1}^m b_j \delta_{y_j} $$
-Our cost function is a matrix $D \in \R_+^{n \times m}$ such that $D_{ij} = d(x_i, y_j)^p$.
-Our joint distribution $P$ is also a matrix of $\R_+^{n \times m}$ that has to satisfy several constraints that can be represented matricially.
-The set which $P$ belongs to is:
 
-$$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbbm{1}_m = a, \, P^\trans \mathbbm{1}_n = b \right\} $$
+The cost function is a matrix $D \in \R_+^{n \times m}$ such that $D_{ij}
+= d(x_i, y_j)^p$. Our joint distribution $P$ is also a matrix of $\R_+^{n
+\times m}$ that has to satisfy several constraints that can be expressed
+matricially: if we denote by $a$ and $b$ the vector of the coefficients of
+the Dirac decompositions of, the set which $P$ has to belong to is:
+
+$$ U(a, b) = \left\{ P \in \R_+^{n \times m}~|~P \mathbf{1}_m = a, \, P^\trans \mathbf{1}_n = b \right\} $$
 
@@
 
-\paragraph{Dual} We now focus on the dual problem. For this we introduce the lagrangian.
+\paragraph{Dual} We now focus on the dual form of this optimization
+problem. For this we introduce the Lagrangian, using the Lagrange multipliers
+$\alpha$ and $\beta$
 
-$$ L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
-The objective function of the dual is:
-$$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle $$
-If the matrix $D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans$ has a negative coefficient, then by having this coefficient grow toward $+\infty$ we obtain a minimum of $-\infty$.
-But if all the coefficients are non negative, then  $\langle P, D - \alpha
-\mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \rangle$ is non negative and we just have to take $P=0$ to obtain a non zero value. We therefore obtain:
+$$
+    L(P, \alpha, \beta) = \langle P, D \rangle + \alpha^\trans (a - P \mathbf{1}_m) + \beta^\trans (b - P^\trans \mathbf{1}_n)
+$$
+
+As usual in a Lagrangian formulation, the objective function of the dual
+problem is:
+
+$$ g(\alpha, \beta) = \min_{P \geqslant 0} L(P, \alpha, \beta) = \min_{P \geqslant 0} \alpha^\trans a + \beta^\trans b + \langle P, D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \rangle $$
+
+If the matrix $D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans$
+has a negative coefficient, then by having the corresponding coefficient
+of $P$ grow toward $+\infty$ we obtain a minimum of $-\infty$. But
+if all the coefficients are non negative, then  $\langle P, D - \alpha
+\mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \rangle$ is non negative
+and we just have to take $P=0$ to obtain a non zero value. We therefore obtain:
+
 $$ g(\alpha, \beta) = \left\{ \begin{array}{ll}
-\alpha^\trans a + \beta^\trans b & \text{si } D - \alpha \mathbbm{1}_m^\trans - \mathbbm{1}_n \beta^\trans \geqslant 0 \\
+\alpha^\trans a + \beta^\trans b & \text{si } D - \alpha \mathbf{1}_m^\trans - \mathbf{1}_n \beta^\trans \geqslant 0 \\
 -\infty & \text{sinon}
@@
 
-Using a minimum flow solver, we can solve this linear optimization problem in
-$\mathcal{O}(n^3 \log (n))$. But the solution is not stable, as illustrated
-on the following figure. If we slightly move our distance $D$ or if we change
-the source and destination mesures, the value of $P$ can drastically change,
-since $P$ is a vertex of our simplex. One solution is to regularize the problem to fall back on a set $U_\alpha(a, b)$
-strictly convex. This set is strictly contained in $U(a, b)$ and the optimal solution of this new problem won't be optimal for the initial problem, but will remain valid.
-Finally, the regularization that we will describe allows to obtain an
-interative algorithm of small complexity.
+We have reduced our initial problem into a linear programming
+formulation. Using a minimum flow solver, we can solve this linear optimization
+problem in $\mathcal{O}(n^3 \log (n))$. But the solution is not stable,
+as illustrated on the following figure. If we slightly modify our distance
+$D$ or if we change the source and destination measures, the value of $P$
+can drastically change, since $P$ is a vertex of our simplex. One solution
+is to regularize the problem to fall back on optimizing in a set $U_\alpha(a,
+b)$ strictly convex. This set is strictly contained in $U(a, b)$ and the
+optimal solution of this new problem will not necessarily be optimal for the
+initial problem, but will remain valid.
+
+We will introduce in the next section a regularization of the original
+problem, that will us to exhibit an iterative algorithm that converges toward
+a meaningful approximation of the minimizer, as well as mainly relying on
+matrix-product kind of linear algebrea and thus potentially implemented
+on GPUs.
+
 \begin{center}
@@
 
-\subs{Regularization}
+\subs{Entropic Regularization}
 
-The regularization that we choose consist in maximizing the entropy of the
-joint-probability $P$. We exhibit a new parameter $\gamma$ that controls
-the regularization and we obtain the following distance:
+The regularization that we highlight consists in solving the optimal
+transport problem, while maximizing the entropy of the joint-probability
+$P$. We therefore exhibit a new parameter $\gamma$ that parametrizes the
+regularization and we obtain the following distance.
 
@@
 \paragraph{Remark}
-Recall that we can consider two random variables $X$ and $Y$ such that $X \sim \mu$, $Y \sim \nu$ and $X,Y \sim P$.
-The entropy of $P$ is maximum when $X$ and $Y$ are independant, \emph{ie.} when $P = ab^\trans$ and this maximum entropy is thus $H(\mu) + H(\nu)$.
-The regularization therefore gives a matrix whose positive values are more
-distributed.
+Recall that we can consider two random variables $X$ and $Y$ such that $X \sim
+\mu$, $Y \sim \nu$ and consider their joint distribution $P$ such that $(X,Y)
+\sim P$. The entropy of $P$ is maximal when $X$ and $Y$ are independant,
+\emph{i.e.} when $P$ factorizes as $P = ab^\trans$ and this maximum entropy
+is given by $H(\mu) + H(\nu)$. The regularization therefore produces a
+matrix whose positive values are more distributed. In other words, we favor
+randomness of the transport map against determinism. The intuition behind
+this regularization comes from the observation that actual trafic patterns
+in networks are sparser than the theoretical solution obtained through
+a minimizer of the Kantorovich problem.
 
@@
 
-By strict convexity, there exist a unique matrix $P_\gamma$ that minimizes the distance:
+By a strict convexity argument, there exist a unique matrix $P_\gamma$
+minimizing the distance:
 $$ P_\gamma = \argmin_{P \in U(a, b)} \langle P, D \rangle - \gamma H(P) $$
@@
 
-We write the new laplacian:
+We write the Laplacian of our new optimization problem.
 
-$$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbbm{1}_m) + \beta^\trans (b - P^\trans \mathbbm{1}_n) $$
-We compute the partial derivative with respect to $P_{ij}$:
-$$ \dfrac{\partial L}{\partial P_{ij}} = M_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$
-The dual is the maximization of the minimum of the laplacian with respect
-to $P$. We must look for $P$ such that the partial derivative is equal to zero:
+$$ L(P, \alpha, \beta) = \sum_{ij} \left( P_{ij} D_{ij} + \gamma P_{ij} \left( \log_2 P_{ij} - 1 \right) \right) + \alpha^\trans (a - P \mathbf{1}_m) + \beta^\trans (b - P^\trans \mathbf{1}_n) $$
+
+We compute the partial derivative with respect to each $P_{ij}$
+$$ \dfrac{\partial L}{\partial P_{ij}} = D_{ij} + \gamma \log_2 P_{ij} - \alpha_i - \beta_j $$
+
+The dual problem is the maximization of the minimum of the Laplacian with
+respect to $P$. We must look for $P$ such that the partial derivative is
+equal to zero:
+
 $$ \dfrac{\partial L}{\partial P_{ij}} = 0 \Rightarrow P_{ij} = e^{\frac{\alpha_i}{\gamma}} e^{-\frac{D_{ij}}{\gamma}}  e^{\frac{\beta_j}{\gamma}} = u_i K_{ij} v_j $$
+
 \findem
@@
 \paragraph{Sinkhorn}
-We have at our disposition an interative method to find this matrix $P_\gamma$. We use the condition $P_\gamma \in U(a,b)$:
+We can derive an iterative method to find this matrix $P_\gamma$. We use
+the condition $P_\gamma \in U(a,b)$:
+
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
@@
 \end{array} \right. $$
+
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
@@
 \end{array} \right. $$
-Then, we introduce $\odot$ the coefficient-wise product:
+
+Then, by introducing $\odot$ the coefficient-wise product, we can rewrite
+this as
+
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrr}
@@
 \end{array} \right. $$
+
+We end up considering a problem which was already studied in the numerical
+analysis community, and known as the \emph{matrix scaling problem}. From
+this formulation, rewritten as below using the coefficient-wise division
+$\oslash$, we can obtain an iterative algorithm performing Sinkhorn's updates.
+
 $$ P_\gamma \in U(a, b) \Leftrightarrow \left\{ \begin{array}{lrc}
-u & = & a \, / \, K v \\
-v & = & b \, / \, K^\trans u
+u & = & a \oslash K v \\
+v & = & b \oslash K^\trans u
 \end{array} \right. $$
-Sinkhorn's algorithm is thus the following: \\
+
+First we modify $u$ such that it satisfies the first equation, and we then
+use this value to modify the value of $v$ in order for it to satisfy the
+second. And we keep performing those updates until convergence. Sinkhorn's
+algorithm is thus the following: \\
+
 \begin{algorithm}
@@
     \Repeat{convergence of $u, v$}{
-		$u \gets a \, / \, K v$ \;
-		$v \gets b \, / \, K^\trans u$ \;
+        $u \gets a \oslash K v$ \;
+        $v \gets b \oslash K^\trans u$ \;
     }
@@
 \paragraph{Complexity}
-It has been proved that the algorithm converges in linear time. Moreover,
-an iteration is done in $\mathcal{O}(mn)$ but the computation can be
-parallelized. It is also possible to use grid convolution to obtain a
-complexity of $\mathcal{O}(n \log n)$.
+It has been proved that the algorithm is a linear convergence scheme. Moreover,
+an iteration is naively done in $\mathcal{O}(mn)$ but the computation can
+easily be parallelized, and especially performed on GPUs, the bottleneck of one
+update being the two matrix products $Kv$ and $K^\trans u$. It is also
+possible to use grid convolution to obtain a complexity of $\mathcal{O}(n
+\log n)$ for one update. More practical precise and exhaustive details can
+be found in Gabriel Peyré and Marco Cuturi's \emph{Computational Optimal
+Transport} \cite{Peyr2018ComputationalOT}.
 
 \paragraph{Dual}
-On peut aussi relever la formulation du dual avec régularisation :
+
+One can also notice the formulation of the dual thanks to the Proposition 1.
 $$ W_\gamma(\mu, \nu) = \max_{\alpha, \beta} \alpha^\trans a + \beta^\trans b - \gamma \left( e^{\alpha / \gamma} \right)^\trans K \left( e^{\beta / \gamma} \right) $$
 
-\paragraph{Lien avec la divergence KL}
-We write the kernel:
+\paragraph{Link with the KL divergence}
+Writing the kernel:
+
 $$ K_\gamma(i, j) = \exp \left( - \dfrac{d_{i, j}}{\gamma} \right) $$
-We obtain a new formulation of the entropic regularization with the KL-divergence:
+
+We obtain a new formulation of the entropic regularized problem using the
+KL-divergence
+
 $$ W_\gamma(\mu, \nu) = \min_{P \in U(a, b)} \gamma KL(P | K_\gamma) $$
@@
 Now that we have convinced ourselves of the good mathematical foundations
-of Optimal Transport, and of the fact that we can actually compute it, we
-can wonder how we can put this tool to great use in the setting of machine
-learning.
+of Optimal Transport, and of the fact that we can actually compute it to
+some extent, we can wonder how we can put this tool to use in the setting
+of machine learning.
 
-\subs{Compute the optimal transport of what?}
+\subs{What do we want to optimally transport ?}
 
-In the previous section we have seen that we have an algorithm to compute
-optimal transport (Sinkhorn's algorithm), and that this algorithm can be
-parallelized. We can take advantage of this fact to actually compute optimal
-transport among huge databases. One of the application of this is similarity
-based retrieval.
+In the previous section we have seen that we have at our disposal an
+algorithm to compute optimal transport maps (Sinkhorn's algorithm), and that
+this algorithm can be parallelized. We can take advantage of this fact to
+actually compute optimal transport among huge databases. Since computing
+optimal transport maps implies that we can compute a meaningful distance
+between two objects, as long as they can be interpreted as distributions,
+one of the application of this is similarity based retrieval.
 
@@
 
-The color-histogram of an image is a tool that characterizes the color
-distribution of an image. Interestingly enough, it has been noticed that
-color-histogram are good tool to describe the similarity of two images. Since
-we now have at our disposal a tool to compute the distance between two
-distribution probability, we can apply it for histograms (that are really not
-that far from probability distribution, so much so that the algorithms and
-methods that we described earlier are relevant even without any fondamental
-change of setting). For a given image we can thus compute the distance between
-its histogram and the histogram of images comming from a bigger database:
-picking images that have the most similar histogram (in the sense of the
-Wasserstein distance) will therefore allow some form of similarity based
+The color-histogram of an image is a tool that characterizes the
+distribution of colors of an image. Interestingly enough, it has been noticed
+that color-histograms are a good tool to describe the similarity of two
+images. Since we now have a tool to compute the distance between two
+distribution probability, we can apply it to histograms (that are essentially
+discretized distribution, so much so that the algorithms and methods that
+we described earlier are relevant even without any fondamental change of
+setting). For a given image we can therefore compute the distance between its
+histogram and the histogram of images comming from a bigger database: picking
+images that have the most similar histogram (in the sense of the Wasserstein
+distance) will therefore allow to perform some form of similarity based
 image retrieval.
@@
     \captionof{figure}{Two images and their respective color-histograms,
-    embedded in $\R^2$}
+    embedded in $\R^2$ (Image from Marco Cuturi's Primer on Optimal Transport at
+    NIPS 2017)}
 \end{center}
@@
 other kind of documents, for example text documents. Some previous work,
-like the $\texttt{word2vec}$ framework allow for embedding of natural
-language words into an Euclidean space. A text therefore defines some sort
-of distribution of probability over this embedded space. As for images we
-can then compare this distribution to the distributions of other texts coming
-from a database (still in the sense of Wasserstein distance), and hopefully
-retrieve text dealing with similar topic, although there may be some
-significant differences in the actual vocabulary of those texts. We take
-advantage of the geometry of the distribution of probability in the embedded
-text, that tries to ensure that similar distributions are coming from texts
-dealing with similar topics.
+such as the $\texttt{word2vec}$ framework allows for the embedding of
+natural language words into an Euclidean space. A text (seen as a collection
+of words) therefore somehow defines a probability distribution over this
+embedded space. As for images we can then compare this distribution to
+the distributions of other texts coming from a database (still in the
+sense of Wasserstein distance), and hopefully retrieve text dealing with
+similar topics, although there may be some significant differences in the
+actual vocabulary of those texts. We can take advantage of the geometry of the
+distribution of probability in the embedded text space, that tries to ensure
+that similar distributions are coming from texts dealing with similar topics.
 
@@
 We can go further than just using the Wasserstein to compute distance between
-distributions and use this as the main tool for the machine learning problems
-that we considered (we previously emphasized retrieval tasks) by introducing
-the Wasserstein distance on well studied machine learning settings and
-problems. We thus shed a new light on those problems, and the results are
-often fruitful. Some authors have coined the term $\textbf{Wassersteinization}$
-for this process.
+distributions and use this notion as distance as a main tool for already
+existing machine learning problems (we previously emphasized retrieval
+tasks for instance) by introducing the Wasserstein distance on well studied
+machine learning settings and problems. We therefore shed a new light on those
+problems, and the results are often fruitful. Some authors have coined the
+term $\textbf{Wassersteinization}$ to describe this process.
 
@@
 
-One way to approach this process is to introduce Wasserstein distance as a
-loss or fidelity term in optimization problems. This will motivate the interest
-toward the study of not only the very Wasserstein distance, but also of its
-derivative. More precisely we can show that we can computationaly evaluate
-the derivatives of the Wasserstein distance. Therefore we can deal with it a variety of optimization problems, including machine lerning related ones.
+One way to approach Wassersteinization is to introduce Wasserstein distance
+as a loss or fidelity term in optimization problems. This will motivate
+the interest toward not only the very Wasserstein distance, but also its
+derivative. More precisely one can show that we can computationaly evaluate
+the derivatives of the Wasserstein distance. Therefore we can deal with it
+in a variety of optimization problems, including machine lerning related ones.
 
@@
 the setting where this point of view as been proposed is the one of data
-averaging. We have already discussed earlier the benefits of Wasserstein's
-barycenters when compared to, say linear interpolations or $l_2$
-barycenters. One can see this put in good use in the case of data acquired
-through real world experiment. For instance, we can imagine a neurology
-setting, where the goal is to study the reaction of the brain of a patient
-to a certain precise stimulus. To do this we subjet the patient to the stimuli
+averaging. We have already discussed earlier the benefits of Wasserstein
+barycenters when compared to, say, linear interpolations or $l^2$
+barycenters. One can put this in good use in the case of data acquired through
+real world experiment. For instance, we can imagine a neuro-imagery setting,
+where the goal is to study the reaction of the brain of a patient to a
+certain precise stimulus. To do this we subjet the patient to the stimuli
 several time and we aim to submit the measures of the brain activity to
-a machine algorithm to study some of its features. But because of a whole
-lot of data acquisition hazards (noise, imprecisions in the measures, etc),
-the result will slighty differ from one experiment to another although it
-concerns the same stimulus. Therefore computing the average of the brain
-activity (that can easily seen to be analog to a probability distribution
-on the brain space), in the sense of Wasserstein barycenters can hopefully
-give access to more relevant data, without loosing the meaningful information.
+a machine algorithm to study some of its features. But because of the huge
+variety of data acquisition hazards (noise, imprecisions in the measures,
+patient variability etc), the result will slighty differ from one experiment
+to another although it concerns the same stimulus. Therefore computing the
+average of the brain activity (that can easily seen to be an analog of a
+probability distribution on the brain space) in the sense of Wasserstein
+barycenters can hopefully give access to more relevant data, without loosing
+the meaningful information.
 
@@
 	\includegraphics[scale=0.3]{brain_map.png}
-    \captionof{figure}{On the left the average brain map among several
-    experiments, on the right the brain activity map on a single experiment.}
+    \captionof{figure}{On the left the standard average brain map among
+    several experiments, on the right the Wasserstein average of the same
+    brain activity data. (Image from \cite{gramfort2015fast}) }
 \end{center}
-
-The same idea can be followed in other fields, like imagery for instance.
 
@@
 huge that it cannot even fit on a single machine, and would it fit that the
-computation time would be out of control. We can imagine to split the dataset
-into $J$ different parts and distribute it amongst $J$ machines running the
-same learning algorithms. The result of this process is a set of $J$
+computation time would be out of reach. We can imagine to split the dataset
+into $J$ different parts and distribute it amongst $J$ machines running
+the same learning algorithms. The result of this process is a set of $J$
 distributions learned from each part of the dataset. Although they represent
-the same truth, because of the variation of the data those variation will
-be slightly different. Once again Wasserstein-averaging provides a useful
-tool to aggregate those distributions, into an hopefully meaningful one,
-closer the underlying real distribution. The good news is that this Wasserstein
+the same truth, because of the variation of the data those distributions will
+be slightly different. Once again Wasserstein-averaging provides a useful tool
+to aggregate those distributions into an hopefully meaningful one, closer
+to the underlying real distribution. The good news is that this Wasserstein
 posterior aggregation method comes with theoretic statistical guaranties!
 
-%\subs{Semi-suppervised learning setting}
+\paragraph{Semi-supervised learning and Wasserstein propagation}
 
-\paragraph{Semi-supervised learning setting}
-
-Wasserstein propagation: smooth a graph $(V,E)$ where an histogram is associated
+We are interested in smoothing a graph $(V,E)$ where an histogram is associated
 $\mu_v$ to each vertex $v$, with a susbset $S$ of the vertices that have a
-fixed histogram (known data).
-
-Solve the following optimization problem.
+fixed histogram (known data). To write this more formally, we want to solve
+the following optimization problem:
 
@@
 
-\subs{Wasserstein PCA, Geodesics and Inverse problem}
+That is, we want to propagate the distribution known at certain vertices
+along the graph, with a "Wasserstein manner". It is a formulation of several
+semi-suppervised learning problems, where we only have at our disposal the
+structure of the graph, and information a small portion of the vertices. The
+distribution that we are handling here can for instance be probability
+distribution of belonging to each class, in a classification problem (which
+leads us to solving the "stereotypical" semi-suppervised problem that is
+vertice labelling), but we could consider broader classes of problems with
+this formalism.
 
-Since the space of distributions with the OT geometry has negative curvature,
-we generalize the PCA (ie. projecting on one dimension spaces) into Generalized
-Principal Geodesics. We then want to explain stuff out of what we found;
-eg. affaire Bonneel vs nounours
+\subs{Wasserstein PCA and Geodesics}
+
+The Wasserstein distance gives the space of probability measure a structure
+of geodesic metric space. Moreover, it has been shown that when $\Omega$
+is an Hilbert space, then the space of probability measures endowed
+with the Wasserstein distance has a Riemannian manifold structure (see
+\cite{ambrosio2008gradient}). Some authors such as \cite{seguy2015principal}
+have proposed to take advantage of this structure to define the notion of
+\textbf{Principal Geodesic Analysis} over this space, which aims to be an
+analog of the good old Principal Component Analysis with Euclidean metrics.
+
+The idea is that, as Ambrosio et al. shown in \cite{ambrosio2008gradient},
+we can characterize the geodesic path between two distributions, that is
+the shortest path with respect to the Wasserstein distance. We denote
+$(\rho_{\mu_1 \to \mu_2}^t)_t$ the geodesic path between $\mu_1$ and $\mu_2$,
+parametrized by $t$. Now, given a family of measures $\nu_1, \nu_2, \dots,
+\nu_N$, we would like to compute the pricipal geodesics of this family, that
+is find a set of geodesics between some measures (that also are to compute)
+that passes through the Wasserstein iso-barycenter of the $\nu_i$ and that
+are close to each of the $\nu_i$.
+
+\begin{center}
+    \begin{tikzpicture}[scale=0.7]
+        % the bottom left border of the surface
+        \path[name path=border1] (0,0) to[out=-10,in=150] (6,-2);
+        % the upper right border of the surface
+        \path[name path=border2] (12,1) to[out=150,in=-10] (5.5,3.2);
+        % a path for a line crossing both borders
+        \path[name path=redline] (0,-0.4) -- (12,1.5);
+        % intersections between the borders and the lines
+        \path[name intersections={of=border1 and redline,by={a}}];
+        \path[name intersections={of=border2 and redline,by={b}}];
+
+        % we draw the surface
+        \shade[left color=whiteGray!90,right color=darkWhite!90] 
+          (0,0) to[out=-10,in=150] (6,-2) -- (12,1) to[out=150,in=-10] (5.5,3.7) -- cycle;
+
+        \coordinate (mu1) at (3.5,0.5);
+        \coordinate (mu2) at (9,1);
+
+        \draw[blue, fill=blue] (mu1) circle (2pt);
+        \draw[blue] (mu1) node[left] {$\mu_1$};
+        \draw[blue, fill=blue] (mu2) circle (2pt);
+        \draw[blue] (mu2) node[right] {$\mu_2$};
+
+        \draw [name path=geodesic, blue] (mu1) to[out=-20, in=-150] node[below] {$\left(\rho_{\mu_1 \to \mu_2}^t\right)_t$} coordinate[pos=0.093] (geo1) coordinate[pos=0.2] (geo2) coordinate[pos=0.53] (geo3) coordinate[pos=0.85] (geo4) (mu2) ;
+
+        \coordinate (nu1) at (5,1.5);
+        \coordinate (nu2) at (6.5,0.4);
+        \coordinate (nu3) at (8,-0.5);
+        \coordinate (nu4) at (4,0.2);
+
+        \draw[red, dashed] (nu4) to (geo1);
+        \draw[red, dashed] (nu1) to[out=-80, in=60] (geo2);
+        \draw[red, dashed] (nu2) to[out=-80, in=70] (geo3);
+        \draw[red, dashed] (nu3) to[out=70, in=-80] (geo4);
+        \draw[red, fill=red] (nu1) circle (2pt) node[above] {$\nu_1$};
+        \draw[red, fill=red] (nu2) circle (2pt) node[above] {$\nu_2$};
+        \draw[red, fill=red] (nu3) circle (2pt) node[below] {$\nu_3$};
+        \draw[red, fill=red] (nu4) circle (2pt) node[below] {$\nu_4$};
+
+    \end{tikzpicture}
+\end{center}
+
+We can formalize this idea more precisely as follows, using for example the
+second order Wasserstein distance $W_2^2$.
+
+$$
+    \min_{\mu_1, \mu_2 \in \mathcal{P}(\Omega)}
+        \sum_{i=1}^N \min_t W_2^2\left(\rho_{\mu_1 \to \mu_2}^t, \nu_i\right)
+$$
+
+Since a couple of measures $(\mu_1, \mu_2)$ entirely characterizes a
+geodesic path, our optimization is performed over the space of probability
+measures over $\Omega$. This problem can be solved using the methods that
+we presented earlier in our \emph{How to compute Optimal Transport} section,
+and some projected gradient descent because of the overall non-convexity of
+this problem. We can iterate this process to gradually explain more and more
+features of our distributions.
+
+\begin{center}
+    \centering
+	\includegraphics[scale=0.3]{ims/MNIST_PCA.png}
+    \captionof{figure}{Principal Geodesic Components obtained from a subset
+    of $2000$ images of the MNIST database, containing an equal proportion
+    of \texttt{2} and \texttt{4}.(Image from \cite{seguy2015principal}) }
+\end{center}

### 2019-10-27 report/papers.tex
Rework some sentences and notations

@@
 
-In this paper written in 2013, Cuturi proposed one of the most currently used algorithm to compute Wasserstein distances.
-He smoothed the classic optimal transport problem thanks to an entropy-regularization and it leads to a new distance which has a faster solver.
-He also showed that using regularization gives better results on MNIST classification problem than using the classic Wasserstein distance or obviously the Euclidean distance.
-This paper is part of the search for adequate and easily computable distances in machine learning.
-Wasserstein distance is better than Euclidean distance in many cases of machine learning but because the complexity to compute it is $\mathcal{O}(n^3 \log n)$ for $n$ the size of the data, Wasserstein distance is rarely used.
-That's why techniques to compute the Wasserstein distance had to be develop.
+In this paper published in 2013, Marco Cuturi proposed one of the currently
+most used algorithm to compute Wasserstein distances. He smoothed the classic
+optimal transport problem thanks to an entropic regularization and it leads
+to a new notion of distance which has a faster solver. This follows the
+regularization idea the we introduced in the first chapter. He also shows
+in this paper that using such a regularization gives better results on
+MNIST classification problems than using the standard Wasserstein distance
+(or obviously using the Euclidean distance). This paper is part of the search
+for adequate and tractable distances in machine learning. Wasserstein distance
+is more suited than Euclidean distance to many problems in machine learning,
+but because the complexity to compute it is $\mathcal{O}(n^3 \log n)$ when
+$n$ is the size of the data, Wasserstein distance is rarely used. That is
+why such techniques to compute the Wasserstein distance had to be developped.
 
 \paragraph{Idea}
-The main ideas of this paper are the use of an entropic-regularization to simplify computation when using Sinkhorn iterative algorithm and the idea of computing the distances between one point and a family of other points in the same time, to use matrix-matrix multiplications instead of matrix-vector multiplications while computing distances between pair of points.
-Thus the algorithm can be implemented on GPU architectures.
+The main ideas of this paper are the use of an entropic-regularization to
+simplify computation when using Sinkhorn iterative algorithm and the idea
+of computing the distances between one point and a family of other points in
+the same time, to use matrix-matrix multiplications instead of matrix-vector
+multiplications while computing distances between pair of points. Thus the
+algorithm can be implemented on GPU architectures.
 
 \paragraph{Regularization}
-We recall that the goal of the optimal transport problem in the Kantorovich formulation is to find a joint distribution $P \in \R^{n \times n}$ where the marginals equal the two distributions for which we want to compute the distance.
-That is to say $P \mathbbm{1}_n = a$ and $P^\trans \mathbbm{1}_n = b$ when we compute the distance between $a$ and $b$.
-We call $U(a, b)$ the set of all possible values for $P$.
-By denoting the entropy by $H$ we can show that:
+We recall that the goal of the optimal transport problem in the Kantorovich
+formulation is to find a joint distribution $P \in \R^{n \times n}$ where
+the marginals equal the two distributions for which we want to compute the
+distance. That is to say $P \mathbf{1}_n = a$ and $P^\trans \mathbf{1}_n = b$
+when we compute the distance between $a$ and $b$. Following the notations
+introduced in the first chapter, we call $U(a, b)$ the set of all possible
+values for $P$. By denoting the entropy by $H$ we can show that:
+
 $$ H(P) \leqslant H(a) + H(b) $$
-And $H(P) = H(a) + H(b)$ when $P = a b^\trans$. We then restrict $P$ to the new set:
-$$ U_\alpha(a, b) = \{ P \in U(a, b) \, | \, \text{KL}(P || a b^\trans) \leqslant \alpha \} = \{ P \in U(a, b) \, | \, H(P) \geqslant H(a) + H(b) - \alpha \} $$
-Two reasons are given to this restriction.
-The first one is that it will be easier to compute the distance.
-The second is that without regularization $P$ has almost $2n-1$ non-zero coefficients.
-Thus the transport is almost deterministic and it is not natural.
-That's why smoothing the transport plan with entropic regularization is presented as a good idea.
-In the paper, it is shown that this restriction induce a new distance:
-$$ d_{M, \alpha} = \min_{P \in U_\alpha(a, b)} \langle P, M \rangle $$
-Were $M$ is a distance matrix.
+
+And $H(P) = H(a) + H(b)$ when $P = a b^\trans$ (that is, when $a$ and $b$ are independant).
+We then restrict $P$ to the new set:
+
+$$ U_\alpha(a, b) = \{ P \in U(a, b)
+\, | \, \text{KL}(P || a b^\trans) \leqslant \alpha \} = \{ P \in U(a, b)
+\, | \, H(P) \geqslant H(a) + H(b) - \alpha \} $$
+
+Two reasons are given to this restriction. The first one is that it will be
+easier to compute the distance. The second is that without regularization
+$P$ has almost $2n-1$ non-zero coefficients. Thus the transport is almost
+deterministic and it is not natural. That is why smoothing the transport
+plan with entropic regularization is presented as a good idea. In the paper,
+it is shown that this restriction induce a new distance:
+
+$$ d_{M, \alpha} = \min_{P \in U_\alpha(a,b)} \langle P, M \rangle $$
+
+Where $M$ is a distance matrix.
 
 \paragraph{Computation}
-By duality theory he obtained that to $\alpha$, corresponds a value $\lambda \geqslant 0$ such that the distance $d_{M, \alpha}$ is equal to the distance $d_M^\lambda$ defined by:
+By duality theory it is obtained that to $\alpha$, corresponds a value
+$\lambda \geqslant 0$ such that the distance $d_{M, \alpha}$ is equal to
+the distance $d_M^\lambda$ defined by:
 $$ d_M^\lambda(a, b) = \langle P^\lambda, M \rangle, \quad \text{where } P^\lambda = \argmin_{P \in U(a, b)} \langle P, M \rangle - \dfrac{1}{\lambda} H(P) $$
-Then Cuturi show that this new distance can be computed with a much cheaper cost than the original Wassertein distance $d_M$.
-In fact the iterative Sinkhorn algorithm presented below converge in few steps to the transport plan $P^\lambda$.
-The \figurename~\ref{recap1} summarize the relationships between all these distances.
+
+Then Cuturi shows that this new distance can be computed with a much cheaper
+cost than the original Wassertein distance $d_M$. In fact the iterative
+Sinkhorn algorithm presented below converge in few steps to the transport
+plan $P^\lambda$. The \figurename~\ref{recap1} summarize the relationships
+between all these distances.
 
 \vspace{3mm}
+
 \begin{algorithm}[H]
@@
 \paragraph{Results}
-The regularized Wasserstein distance is then used in a SVM to classifiy images of the MNIST database.
-The \figurename~\ref{comp} compare the error on a test set for different distances used in the SVM.
-The distance EMD is the classical optimal transport.
-As we can see the regularized version beat all distances.
-Furthermore he evaluated the execution time by stopping the iterations when $\| d_{i+1} / d_i - 1 \| < 10^{-4}$ where $d_i$ is the distance obtained at the end of the iteration $i$.
-He gave a comparison between EMD and regularized Wasserstein for different values of $\lambda$ in the \figurename~\ref{comp2}.
+The regularized Wasserstein distance is then used in a SVM to classifiy
+images of the MNIST database. The \figurename~\ref{comp} compare the error
+on a test set for different distances used in the SVM. The distance EMD
+is the classical optimal transport. As we can see the regularized version
+beats all other distances. Furthermore he evaluated the execution time by
+stopping the iterations when $\| d_{i+1} / d_i - 1 \| < 10^{-4}$ where $d_i$
+is the distance obtained at the end of the iteration $i$. He gave a comparison
+between EMD and regularized Wasserstein for different values of $\lambda$
+in the \figurename~\ref{comp2}, highlighting the efficiency of computing
+the regularized optimal transport.
 
@@
 
-This paper proposes a new GAN training algorithm based on Wasserstein distance.
-When we learn a probability we need to use a divergence between probability distributions.
-In a first time they show the advantages of the Wasserstein distance over all well-known other divergences on probabilities.
-Thanks to their WGAN (for Wasserstein GAN), they cure the problem of the need of maintaining a balance between training discriminator and training generator.
-They also observed that the mode dropping phenomenon is reduced.
-Furthermore their GAN allows to plot the Wasserstein distance continuously which is very usefull for debugging and to have an estimation of the quality.
+This paper proposes a new GAN training algorithm based on Wasserstein
+distance. When we learn a probability distribution we need to have a notion of
+divergence between probability distributions. In a first time they show the
+advantages of the Wasserstein distance over all well-known other divergences
+on probabilities. Thanks to their WGAN (for Wasserstein GAN), they cure the
+problem of the need of maintaining a balance between training discriminator
+and training generator. They also observed that the mode dropping phenomenon
+is reduced. Furthermore their GAN allows to plot the Wasserstein distance
+continuously which is very usefull for debugging and having an estimation
+of the quality.
 
 \paragraph{Advantage of Wasserstein distance}
-They illustrate the interest of Wasserstein distance on the convergence of the probability $\Pp_\theta = g_\theta(Z) = (\theta, Z)$ toward $\Pp_0 = (0, Z)$ for $Z \sim U([0, 1])$ when $\theta$ tends to 0.
-We have:
+They illustrate the interest of Wasserstein distance with a toy problem,
+that is the convergence of the probability $\Pp_\theta = g_\theta(Z) =
+(\theta, Z)$ toward $\Pp_0 = (0, Z)$ for $Z \sim U([0, 1])$ when $\theta$
+tends to 0. We have:
 \begin{itemize}
@@
 \end{itemize}
-Thus the Wasserstein distance is the only one for which there is convergence.
-Then a gradient descent can be applied with the Wasserstein distance but it can't be applied with other divergences.
-This is a just an example but they also proved the following theorem which shows the continuity under some assumptions:
+
+Furthermore the Wasserstein distance is the only one for which there is
+convergence. Then a gradient descent can be applied using the Wasserstein
+distance while it cannot be applied with other divergences. This is a just an
+example but they also proved the following theorem which shows the continuity
+under some assumptions.
 
@@
 
-They also give another theorem that well show that the Wasserstein distance is a very natural divergence for probability distributions:
+They also give another theorem that well show that the Wasserstein distance
+is a very natural divergence for probability distributions.
 
 \PROP{
-	Consider a sequence of distributions $(\Pp_n)$ on a compact space $\mathcal{X}$.
-If $W(\Pp_n, \Pp) \rightarrow 0$ for a certain distribution $\Pp$, then $\Pp_n \xrightarrow{\mathcal{D}} \Pp$ where the limit is in the sense of the convergence in distribution.
+	Consider a sequence of distributions $(\Pp_n)$ on a compact space
+	$\mathcal{X}$. If $W(\Pp_n, \Pp) \rightarrow 0$ for a certain
+	distribution $\Pp$, then $\Pp_n \xrightarrow{\mathcal{D}} \Pp$
+	where the limit is in the sense of the convergence in distribution.
 }
@@
 \paragraph{WGAN}
-To construct their GAN they need to compute the gradient of $W(\Pp_r, \Pp_\theta)$ with respect to $\theta$. For that they consider the following dual formulation of the Wasserstein distance:
+To construct their GAN they need to compute the gradient of $W(\Pp_r,
+\Pp_\theta)$ with respect to $\theta$. For that they consider the following
+dual formulation of the Wasserstein distance:
+
 $$ W(\Pp_r ,\Pp_\theta) = \max_{\| f \|_L \leqslant 1} \E_{x \sim \Pp_r} \left[ f(x) \right] - \E_{x \sim \Pp_\theta} \left[ f(x) \right] $$
-Where $\| f \|_L$ denote the Lipshitz coefficient of $f$. If $f$ is the solution to the maximization problem then the gradient is:
+
+Where $\| f \|_L$ denote the Lipshitz coefficient of $f$. If $f$ is the
+solution to the maximization problem then the gradient is:
+
 $$ \nabla_\theta W(\Pp_r, \Pp_\theta) = - \E_{z \sim Z} \left[ \nabla_\theta f(g_\theta(z)) \right] $$
-To find $f$ they train a neural network parameterized with weights $w$ in a compact space $\mathcal{W}$ that lead to a function $f_w$.
-Then they can backprop $\theta$ using $\E_{z \sim Z} \left[ \nabla_\theta f_w (g_\theta(z)) \right]$.
-To have $w$ in a compact space, they use weight clamping.
-Their algorithm is in the \figurename~\ref{algo}.
+
+To find $f$ they train a neural network parameterized with weights $w$
+in a compact space $\mathcal{W}$ that lead to a function $f_w$. Then they
+can backpropagate $\theta$ using $\E_{z \sim Z} \left[ \nabla_\theta f_w
+(g_\theta(z)) \right]$.  To have $w$ in a compact space, they use weight
+clamping. Their algorithm is in the \figurename~\ref{algo}.
 
@@
 
-\paragraph{results}
-First, the authors set up a small experiment to showcase the difference between GAN and WGAN (\figurename~\ref{compa}).
-There are two 1D Gaussian distributions, blue for real and green for fake.
-Train a GAN discriminator and WGAN critic to optimality, then plot their values over the space.
-The red curve is the GAN discriminator output, and the cyan curve is the WGAN critic output.
+\paragraph{Results}
+First, the authors set up a small experiment to showcase the difference
+between GAN and WGAN (\figurename~\ref{compa}). There are two 1D Gaussian
+distributions, blue for real and green for fake. Train a GAN discriminator and
+WGAN critic to optimality, then plot their values over the space. The red curve
+is the GAN discriminator output, and the cyan curve is the WGAN critic output.
 
@@
 
-Both identify which distribution is real and which is fake, but the GAN discriminator does so in a way that makes gradients vanish over most of the space.
-In contrast, the weight clamping in WGAN gives a reasonably nice gradient over everything.
+Both identify which distribution is real and which is fake, but the GAN
+discriminator does so in a way that makes gradients vanish over most of
+the space. In contrast, the weight clamping in WGAN gives a reasonably nice
+gradient over the whole space.
 
-Another important point is that the quality of the generation is correlated to Wasserstein estimation as we can see in the \figurename~\ref{meaning}.
-We  can also see that the training of the critic function $f$ works well as the Wasserstein estimator strictly decrease.
-The gradient of the loss function is well computed.
-But this is not a method to compare the quality of different GANs as the estimator is defined with a constant scaling factor depending on the critic's architecture.
-They finally observed an improvement of the stability.
+Another important point is that the quality of the generation is correlated
+to Wasserstein estimation as we can see in the \figurename~\ref{meaning}. We
+can also see that the training of the critic function $f$ works well as the
+Wasserstein estimator strictly decrease. The gradient of the loss function is
+well computed. But this is not a method to compare the quality of different
+GANs as the estimator is defined with a constant scaling factor depending
+on the critic's architecture. They finally observed an improvement of the
+stability.
 
@@
 
-Because a single step of Sinkhorn algorithm has a complexity in $\mathcal{O}(n^2)$, computing the Wassertein distance may take a while in many cases.
-That's why the researchers who write this paper thought about a more effective solution to compute a step.
-They found a way to use Gaussian convolution with a $\mathcal{O}(n \log n)$ complexity instead of a matrix-vector product.
-More generally they use a heat kernel.
-So instead of the convolution it can also be a sparse pre-factored linear system which can be computed in a complexity smaller than $\mathcal{O}(n^2)$.
-Furthermore by doing this, the convergence is still linear.
-Of course this can not be applied in all cases of optimal transport, but it can be used on very common geometric domains, like images or meshes.
-In addition of the computation of Wasserstein distance, they also propose a way to use convolution in the computation of Wasserstein barycentres and Wasserstein propagation.
+Because a single step of Sinkhorn algorithm has a complexity of
+$\mathcal{O}(n^2)$, computing the Wassertein distance may take a while in
+many cases, even though the number of iterations needed to converge toward
+a good minimizer is small. That is why the researchers who wrote this paper
+thought about a more effective solution to perform a step of the
+algorithm. They found a way to use Gaussian convolution with a $\mathcal{O}(n
+\log n)$ complexity instead of a matrix-vector product. More generally they
+use a heat kernel. So instead of the convolution it can also be a sparse
+pre-factored linear system which can be computed in a complexity smaller
+than $\mathcal{O}(n^2)$. Furthermore this approach does not affect the overall
+convergence speed of the scheme, which remains linear. Of course this can
+not be applied to all the cases of optimal transport, but it can be used
+on very common geometric domains, like images or meshes. In addition of the
+computation of Wasserstein distance, the authors also propose a way to use
+convolution in the computation of Wasserstein barycenters and Wasserstein
+propagation.
 
 \paragraph{Using a heat kernel in the entropy-regularization}
-We recall what is the entropy-regularized Wasserstein distance between $\mu_0$ and $\mu_1$ on the domain $M$:
-$$ W_{2, \gamma}^2(\mu_0, \mu_1) = \inf_{\pi \in \Pi} \left[ \int \int_{M \times M} d(x, y)^2 \pi(x, y) dx dy \, - \, \gamma H(\pi) \right] = \gamma \left[ 1 + \min_{\pi \in \Pi} \text{KL}(\pi | \mathcal{K}_\gamma) \right]$$
-Where:
-$$ \Pi = \left\{ \pi \in \text{Prob}(M \times M) \, | \, \pi(\cdot, M) = \mu_0, \pi(M, \cdot) = \mu_1 \right\} $$
-$$ H(\pi) = - \int \int_{M \times M} \pi(x, y) \ln \pi(x, y) dx dy $$
-$$ \mathcal{K}_\gamma = e^{-d(x, y)^2 / \gamma} $$
-$$ \text{KL}(\pi | \mathcal{K}) = \int \int_{M \times M} \pi(x, y) \left[ \ln \dfrac{\pi(x, y)}{\mathcal{K}(x, y)} - 1 \right] dx dy $$
-This is a strictly convex problem thanks to the entropy.
-The idea of the paper is to use heat kernel because in some cases, the solution of the diffusion equation can be computed in a effective way.
-We denote by $\mathcal{H}_t(x, y)$ the diffusion between $x$ and $y$ after a time $t$ in the heat kernel.
-Just remember that the heat kernel is a solution of the heat equation:
-$$ \dfrac{\partial f}{\partial t} = \Delta f $$
-One can show the following approximation:
+We recall what is the entropy-regularized Wasserstein distance between $\mu_0$
+and $\mu_1$ on the domain $M$:
+
+$$
+    W_{2, \gamma}^2(\mu, \nu)
+        = \inf_{\pi \in \Pi(\mu, \nu)} \left[ \int \int_{M\times M} d(x, y)^2 \pi(x, y) dx dy \, - \, \gamma H(\pi) \right]
+        = \gamma \left[ 1 + \min_{\pi \in \Pi(\mu,\nu)} \text{KL}(\pi | \mathcal{K}_\gamma) \right]
+$$
+
+Where have kept the same notations as the ones used in the first chapter,
+that is
+
+$$
+    \Pi = \left\{ \pi \in \mathcal{P}(M \times M) \mid \pi(\cdot, M) = \mu_0, \pi(M, \cdot) = \mu_1 \right\}
+$$
+
+$$
+    H(\pi) = - \int \int_{M \times M} \pi(x, y) \ln \pi(x, y) dx dy
+$$
+
+$$
+    \mathcal{K}_\gamma = e^{-d(x, y)^2 / \gamma}
+$$
+$$
+    \text{KL}(\pi | \mathcal{K}) = \int \int_{M \times M} \pi(x, y) \left[ \ln \dfrac{\pi(x, y)}{\mathcal{K}(x, y)} - 1 \right] dx dy
+$$
+
+This is a strictly convex problem thanks to the entropy. The idea of the paper
+is to use heat kernel because in some cases, the solution of the diffusion
+equation can be computed in a effective way. We denote by $\mathcal{H}_t(x,
+y)$ the diffusion between $x$ and $y$ after a time $t$ in the heat kernel. Just
+remember that a heat kernel is a solution of the heat equation:
+
+$$
+    \dfrac{\partial f}{\partial t} = \Delta f
+$$
+
+One can show the following that the following approximation holds
+
 $$ \mathcal{K}_\gamma \approx \mathcal{H}_{\gamma / 2}(x, y) $$
-We won't store $\mathcal{H}$ because it could have a space complexity in $\mathcal{O}(n^2)$ and we want a lower complexity.
-But we generally know a way to apply $\mathcal{H}$ to a vector.
-In the case of images we apply Gaussian convolution with $\sigma^2 = \gamma$.
-\\
-In the case of triangle meshes we associate a weight to faces proportional to their area.
-We denote by $a$ the vector of these weights (In images of size $n \times m$ we set $a = 1 / (nm)$).
-Then we denote by $L$ the cotangent Laplacian and by $D_a$ the diagonal matrix with diagonal $a$.
-By discretizing the heat equation, we obtain:
-$$ w = \mathcal{H}_t(v) \Leftrightarrow \left( D_a + t L \right) w = v $$
-The linear system can be solved efficiently by pre-computing a sparse Cholesky factorization.
+
+We won't store $\mathcal{H}$ because it could have a space complexity in
+$\mathcal{O}(n^2)$ and we want a lower complexity. But we generally know
+a way to apply $\mathcal{H}$ to a vector. In the case of images we apply
+Gaussian convolution with $\sigma^2 = \gamma$. In the case of triangle meshes
+we associate a weight to faces proportional to their area. We denote by $a$
+the vector of those weights (for images of size $n \times m$ we set $a = 1 /
+(nm)$). Then we denote by $L$ the cotangent Laplacian and by $D_a$ the diagonal
+matrix whose diagonal is $a$. By discretizing the heat equation, we obtain:
+
+$$
+    w = \mathcal{H}_t(v) \Leftrightarrow \left( D_a + t L \right) w = v
+$$
+
+The linear system can be solved efficiently by pre-computing a sparse
+Cholesky factorization.
 
 \paragraph{Algorithm}
-With some computations we arrive at:
+With some computations we arrive at
 
@@
 
-Where $\oslash$ and $\otimes$ denote element-wise operations.
-As Sinkhorn algorithm we obtain a simple iterative algorithm where, this time, steps can be computed in $\mathcal{O}(n \log n)$ in many cases, where $n$ is the size of the domain $M$.
+Where $\oslash$ and $\otimes$ denote element-wise operations. As with classical
+Sinkhorn algorithm we obtain a simple iterative algorithm where, this time,
+each step can be computed in $\mathcal{O}(n \log n)$ in many cases, $n$
+being the size of the domain $M$.
 
 \paragraph{Barycenter}
-An algorithm for computing barycenters is also provided.
-We won't enter into details but we give their algorithm for computing the barycenter of distributions $\mu_i$ associated with weights $\alpha_i$ for $1 \leqslant i \leqslant k$:
+An algorithm for computing barycenters is also provided. We won't enter
+into details but we give their algorithm for computing the barycenter of
+distributions $\mu_i$ associated with weights $\alpha_i$ for $1 \leqslant
+i \leqslant k$:
 
@@
 \vspace{3mm}
-Where entropic sharpening allow us to make the entropy of the result $\mu$ smaller than the maximum entropy of all $\mu_i$.
-To do so if $H(\mu)$ is greater than $\max_i H(\mu_i)$, then we set $\mu$ to $\mu^\beta$ where $\beta$ is such that $H(\mu^\beta) = \max_i H(\mu_i)$.
-We can find $\beta$ using Newton's method.
+Here entropic sharpening allows us to make the entropy of the result $\mu$
+smaller than the maximum entropy of all $\mu_i$. To do so if $H(\mu)$
+is greater than $\max_i H(\mu_i)$, then we set $\mu$ to $\mu^\beta$ where
+$\beta$ is such that $H(\mu^\beta) = \max_i H(\mu_i)$. We can find such a
+$\beta$ using Newton-like methods.
 
@@
 \end{itemize}
-Finally \figurename~\ref{barp} and \figurename~\ref{cardp} are two images of barycenters they have computed.
-The first one is an image where we find barycenters of four images that are in the corners.
-The second is an interpolation between two cards.
-As we tried to reproduce it in the implementation part we put them here:
+
+Finally \figurename~\ref{barp} and \figurename~\ref{cardp} are two images
+of barycenters that they have computed. The first one is an image where
+we find several barycenters of four images that are in the corners. The
+second is an interpolation between two cards. As we tried to reproduce it
+in our implementation part we put them here, for comparison sakes:
 
